### 2024-04-02 paper/paper.qmd
update

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Yihang  Cai
thanks: "Code and data are available at: https://github.com/peachvegetable/NBA-player-points"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

### work space setup ###
library(tidyverse)
library(arrow)
library(tidymodels)
library(modelsummary)
library(ggplot2)
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

```{r}
predictions <- predict(second_lasso_model, test_data)
results <- bind_cols(test_data, predictions)
# Calculate RMSE
rmse_results_2 <- rmse(results, truth = pts, estimate = .pred)
rsq_results_2 <- rsq(results, truth = pts, estimate = .pred)
```

# Results

Our results are summarized in @tbl-modelresults.

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

```{r}
analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")

predictors <- c("g", "gs", "mp", "fg_percent", "x3p_percent", 
                "x2p_percent", "e_fg_percent", "ft_percent", 
                "orb", "drb", "trb", "ast", "stl", "blk", "tov", "pf")

for (predictor in predictors) {
  p <- ggplot(analysis_data, aes_string(x = predictor, y = "pts")) + 
    geom_point(alpha = 0.5) + 
    geom_smooth(method = "lm", se = FALSE, color = "blue") + 
    ggtitle(paste("Points vs", predictor)) +
    theme_minimal()

  print(p)
}
```

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

\newpage

# References

### 2024-04-02 scripts/04-model.R

@@ -13,7 +13,7 @@ library(tidymodels)
 library(arrow)
 
 #### Read data ####
-analysis_data <- read_parquet("data/analysis_data/analysis_data_1.parquet")
+analysis_data <- read_parquet("data/analysis_data/analysis_data.parquet")
 
 ### Model data ####
 
@@ -62,19 +62,9 @@ final_lasso_workflow <- finalize_workflow(
 # fit the model
 first_lasso_model <- fit(final_lasso_workflow, data = train_data)
 
-predictions <- predict(first_lasso_model, test_data)
-results <- bind_cols(test_data, predictions)
-# Calculate RMSE
-rmse_results <- rmse(results, truth = pts, estimate = .pred)
-rsq_results <- rsq(results, truth = pts, estimate = .pred)
+# Second model with feature engineering
 
-
-
-
-
-
-
-analysis_data_1 <- read_parquet("data/analysis_data/analysis_data_1.parquet")
+analysis_data_1 <- read_parquet("data/analysis_data/analysis_data.parquet")
 analysis_data_1$age_squared = analysis_data_1$age^2
 analysis_data_1$pts_per_min = analysis_data_1$pts / analysis_data_1$mp
 analysis_data_1$ast_per_min = analysis_data_1$ast / analysis_data_1$mp
@@ -126,17 +116,16 @@ final_lasso_workflow <- finalize_workflow(
 # fit the model
 second_lasso_model <- fit(final_lasso_workflow, data = train_data)
 
-predictions <- predict(second_lasso_model, test_data)
-results <- bind_cols(test_data, predictions)
-# Calculate RMSE
-rmse_results_2 <- rmse(results, truth = pts, estimate = .pred)
-rsq_results_2 <- rsq(results, truth = pts, estimate = .pred)
-
 
 #### Save model ####
 saveRDS(
   first_lasso_model,
   file = "models/first_lasso_model.rds"
+)
+
+saveRDS(
+  second_lasso_model,
+  file = "models/second_lasso_model.rds"
 )
 
 

### 2024-04-02 paper/paper.qmd
update

@@
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+The dataset is obtained from Basketball Reference @citeDataset, which contains NBA players' statistics in the 2023 to 2024 season.
+The dataset is directly downloaded from Basketball Reference using the given instruction.
+The dataset contains different statistics of NBA players, for instance players' position, age, assists, steals, etc.
+There are in total of 718 observations before any data cleanings.
 
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-Talk way more about it.
+Since this dataset contains 3-point field goals, 2-point field goals, free throws and field goals, points can be calculated solely using these variables.
+My goal is to predict a the numebr of scores a NBA player scores based on his performance, which means that these features are too closely related to the target variable (i.e.
+points).
+Thus, those variables are removed from the dataset using 'tidyverse' package in R @citeTidyverse.
+Moreover, there are 12 different positions in the dataset, I categorized them into a broader classification which has only three categories - Guards (G): SG, PG, SG-PG, PG-SG, Forwards (F): SF, PF, PF-SF, SF-PF, SF-SG, Centers (C): C, PF-C, C-PF.
+This is done so that I don't need to create overwhelming dummy variables when generating the model and this process of binning could simplify the model and potentially strengthen the signal by reducing noise and multicollinearity.
 
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+Lasso regression is a variation of linear regressions that can actually perform feature selection by setting the coefficients of less important features to zero.
+It has an additional parameter $\lambda$, which is the regularization parameter that controls the strength of the penalty.
+The penalty term shrinks some of the coefficients (i.e.
+the less important variables) toward zero.
+As $\lambda$ increases, more coefficients are set to zero, leading to a simpler model.
+The value of $\lambda$ is determined through cross-validation.
 
@@
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
 \begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
+y_i = \beta_0 + \beta_i \cdot X_i
 \end{align}
 
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+In this equation, $y_i$ is the number of points a player scores, which is the dependent variable I am trying to predict.
+$\beta_0$ is the interception, and $\beta_i$ is a matrix that contains the coefficients $\beta_1, \beta_2, ..., \beta_{18}$ for each predictor that the lasso regression will estimate.
+$X_i$ is also a matrix contains the predictors: players position, age, games, game starts, minutes played, field goal percentage, 3-point field goal percentage, 2-point field goal percentage, effective field goal percentage, free throw percentage, offensive rebounds, defensive rebounds, total rebounds, assists, steals, blocks, turnovers, and personal fouls.
+
+We run the model in R [@citeR] using the `tidymodels` package of @citeTidymodels.
 
@@
 ```{r}
-predictions <- predict(second_lasso_model, test_data)
-results <- bind_cols(test_data, predictions)
-# Calculate RMSE
-rmse_results_2 <- rmse(results, truth = pts, estimate = .pred)
-rsq_results_2 <- rsq(results, truth = pts, estimate = .pred)
+# #| include: false
+# #| warning: false
+# #| message: false
+# 
+# first_lasso_model <- readRDS("../models/first_lasso_model.rds")
+# second_lasso_model <- readRDS("../models/second_lasso_model.rds")
+# predictions <- predict(second_lasso_model, test_data)
+# results <- bind_cols(test_data, predictions)
+# # Calculate RMSE
+# rmse_results_2 <- rmse(results, truth = pts, estimate = .pred)
+# rsq_results_2 <- rsq(results, truth = pts, estimate = .pred)
 ```
@@
 ```{r}
+#| include: false
+#| warning: false
+#| message: false
+
 analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")

### 2024-04-07 paper/datasheet_template.qmd
datasheet partly finished

---
title: "Datasheet for NBA player stats"
subtitle: "My subtitle if needed"
author: 
  - Yihang Cai
thanks: "Code and data are available at: https://github.com/peachvegetable/NBA-player-points"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The dataset was created to enable detailed analysis and predictive modeling of NBA players' performance for the 2023-2024 season. It aims to predict the number of points a player scores based on various performance metrics.
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - The dataset was created by Basketball Reference, and I downloaded following the instruction given. Utilizing publicly available data from Basketball Reference, which sources its current-season data from SportRadar, the official statistics provider of the NBA.
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - TBD
4. *Any other comments?*
    - TBD

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- Each instance in the dataset represents the season-long aggregated performance statistics for an individual NBA player for the 2023-2024 season. This includes games played, various scoring metrics, rebounding, assists, and other standard basketball statistics.
2. *How many instances are there in total (of each type, if appropriate)?*
	- The dataset contains 718 observations, each corresponding to a unique player or a player's stint with a particular team.
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- The dataset contains a comprehensive set of NBA player statistics for the specified season, rather than a sample. It includes all players who participated in the NBA during the 2023-2024 season, accounting for any who played sufficient minutes to be included in standard statistical aggregations.
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- Personal and Demographic Information: Player's name, position, age, and team.
  - Game Participation: Games played and started, minutes played.
  - Scoring Metrics: Field goals, three-point field goals, and free throws (made, attempted, and percentages).
  - Rebounding: Offensive, defensive, and total rebounds.
  - Playmaking and Defense: Assists, steals, blocks.
  - Game Conduct: Turnovers, personal fouls.
  - Overall Performance: Total points scored.
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- The dataset does not inherently contain a designated label or target for predictive modeling purposes. However, any of the numerical statistics, such as total points scored (PTS), could serve as a target variable for predictive analyses.
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- TBD
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- The dataset does not explicitly encode relationships between instances (players). However, relationships or comparisons could be inferred or constructed based on team affiliations, positions, or other shared attributes.
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- For predictive modeling, such as predicting a player's points based on their performance, a common approach is to split the data into training, validation, and testing sets. A typical split ratio could be 70% for training, 15% for validation, and 15% for testing. The training set is used to train the model, the validation set is for tuning model parameters and preventing overfitting, and the testing set is for evaluating the final model's performance. This split ensures that the model is evaluated on unseen data, providing an unbiased estimate of its performance.
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- The dataset might contain redundancies, especially if it includes separate entries for players who have been traded or played for multiple teams within the season. These instances might need to be consolidated or treated separately, depending on the analysis.
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- The dataset is self-contained in terms of providing a comprehensive set of season-long player statistics. It does not rely on external resources for its primary content, but additional context or data (e.g., team performance, player salaries) might require external sources.
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- The dataset contains publicly available NBA player statistics and does not include any personal or confidential information. Data like player performance metrics are standard in sports analytics and are widely disseminated without privacy concerns.
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- The dataset is purely statistical and related to professional basketball performance. It does not contain any content that could be considered offensive, insulting, threatening, or anxiety-inducing.
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- The dataset may allow for the identification of sub-populations based on age (as age is a recorded attribute) and position played, which might correlate with certain physical attributes or skill sets. However, it does not explicitly categorize players by other demographic variables such as gender, as the dataset pertains solely to an all-male professional sports league.
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- Yes, individuals can be directly identified from the dataset as it contains the names of NBA players. The dataset is intended to provide statistics at the individual player level, making identification not only possible but a fundamental feature of the data.
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- The dataset does not contain sensitive personal information. It is focused on professional performance metrics in the context of NBA basketball games, such as points scored, assists, rebounds, etc. It lacks personal data such as race, ethnic origins, sexual orientations, religious beliefs, political opinions, financial or health data, or any forms of government identification.
16. *Any other comments?*
	- This dataset is valuable for sports analytics purposes, offering insights into player performance and league dynamics. Its content is ethically unproblematic, being limited to professional achievements in a public sporting context.

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- TBD
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- TBD
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- TBD
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- TBD
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- TBD
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- TBD
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- TBD
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- TBD
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- TBD
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- TBD
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- TBD
12. *Any other comments?*
	- TBD

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- TBD
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- TBD
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- TBD
4. *Any other comments?*
	- TBD

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- TBD
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- TBD
3. *What (other) tasks could the dataset be used for?*
	- TBD
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- TBD
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- TBD
6. *Any other comments?*
	- TBD

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- TBD
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- TBD
3. *When will the dataset be distributed?*
	- TBD
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- TBD
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- TBD
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- TBD
7. *Any other comments?*
	- TBD

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- TBD
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- TBD
3. *Is there an erratum? If so, please provide a link or other access point.*
	- TBD
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- TBD
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- TBD
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- TBD
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- TBD
8. *Any other comments?*
	- TBD

\newpage

# References

### 2024-04-08 paper/datasheet_template.qmd
test file finished

@@
 If so, please describe how.*
-	- TBD
+	- The data was directly observable and publicly available, compiled from official NBA games and player performance statistics.
+These statistics are typically recorded by official scorers and statisticians during games and are not reported by the subjects (players) themselves.
+The data is validated through the official NBA data validation processes, which involve real-time and post-game reviews to ensure accuracy.
 2.
@@
 How were these mechanisms or procedures validated?*
-	- TBD
+	- The collection involved both hardware apparatuses (like tracking devices and cameras in arenas) and software programs (for data aggregation and presentation).
+NBA and SportRadar, as the official statistics provider, likely use a combination of these tools to compile and validate game statistics.
+The validation processes are rigorous, involving multiple checks to ensure data accuracy and reliability.
 3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
-	- TBD
+	- This dataset is not a sample but rather a comprehensive compilation of all player statistics for the 2023-2024 NBA season, representing a complete set of data points for that timeframe.
 4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
-	- TBD
+	- The data was collected by official NBA statisticians and game officials, possibly with support from SportRadar personnel.
+These individuals are professionals employed or contracted by the NBA and SportRadar, compensated according to their professional agreements, not crowdworkers.
 5.
@@
 If not, please describe the timeframe in which the data associated with the instances was created.*
-	- TBD
+	- The data was collected over the course of the 2023-2024 NBA season.
+The timeframe of data collection matches the creation timeframe of the data, as it is recorded live during games and updated in near real-time.
 6.
@@
 If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
-	- TBD
+	- Given the public and professional nature of the data, formal ethical review processes like those conducted by institutional review boards are generally not applicable.
+The data collection is part of standard operations within professional sports leagues.
 7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
-	- TBD
+	- The data was obtained via a third party, the Basketball Reference website, which aggregates official NBA statistics.
 8.
@@
 If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
-	- TBD
+	- As professional athletes, NBA players are aware that their performance metrics are recorded and publicly disseminated as part of their participation in the league.
+Formal individual notifications for each data collection instance are not standard practice.
 9.
@@
 If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
-	- TBD
+	- Players consent to the recording and public sharing of their game performance as part of their professional contracts with the NBA and their respective teams.
+Explicit consent for each instance of data collection in the form of game statistics is inherent to their professional engagement.
 10.
@@
 If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
-	- TBD
+	- The public sharing of game performance data is a standard part of professional sports contracts.
+Individual players typically do not have the option to revoke consent for the sharing of standard game performance data.
 11.
@@
 If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
-	- TBD
+	- An analysis of this nature is generally not conducted for publicly available professional sports performance data, as it is considered non-sensitive and is a standard aspect of professional sports analytics.
 12. *Any other comments?*
-	- TBD
+	- The collection and use of this data are standard practices within the domain of professional sports analytics and are not subject to the typical privacy concerns associated with personal or sensitive data.
 

### 2024-04-08 scripts/03-test_data.R

@@ -1,12 +1,85 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
+# Purpose: Tests the processed dataset analysis_data
 # Author: Yihang Cai
 # Date: 30 Mar 2024
 # Contact: Yihang.cai.nz@gmail.com
-# Any other information needed? Some of the codes are adpated from Telling Stories with Data by Rohan Alexander
+# Any other information needed? Some of the codes are adapted from Telling Stories with Data by Rohan Alexander
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(testthat)
+library(arrow)
 
 #### Test data ####
+analysis_data <- read_parquet("data/analysis_data/analysis_data.parquet")
+
+# Test Data Types
+test_that("Data types are correct", {
+  expect_type(analysis_data$player, "character")
+  expect_type(analysis_data$age, "double")
+  expect_type(analysis_data$g, "double")
+  expect_type(analysis_data$gs, "double")
+  expect_type(analysis_data$mp, "double")
+  expect_type(analysis_data$fg_percent, "double")
+  expect_type(analysis_data$x3p_percent, "double")
+  expect_type(analysis_data$x2p_percent, "double")
+  expect_type(analysis_data$e_fg_percent, "double")
+  expect_type(analysis_data$ft_percent, "double")
+  expect_type(analysis_data$orb, "double")
+  expect_type(analysis_data$drb, "double")
+  expect_type(analysis_data$trb, "double")
+  expect_type(analysis_data$ast, "double")
+  expect_type(analysis_data$stl, "double")
+  expect_type(analysis_data$blk, "double")
+  expect_type(analysis_data$tov, "double")
+  expect_type(analysis_data$pf, "double")
+  expect_type(analysis_data$pts, "double")
+  expect_type(analysis_data$broad_position, "character")
+})
+
+# Test Max/Min Values for each variable
+test_that("Games_Played, Points, and Assists are within expected ranges", {
+  expect_true(all(analysis_data$g >= 1 & analysis_data$gs >= 0))
+  expect_true(all(analysis_data$pts >= 0))
+  expect_true(all(analysis_data$ast >= 0))
+})
+
+# Test Total Number of Observations
+test_that("Total number of observations is correct", {
+  expect_equal(nrow(analysis_data), 616)
+})
+
[…]

### 2024-04-08 paper/paper.qmd
update

@@
 library(ggplot2)
+library(knitr)
+library(janitor)
 ```
@@
 The dataset is directly downloaded from Basketball Reference using the given instruction.
-The dataset contains different statistics of NBA players, for instance players' position, age, assists, steals, etc.
+The dataset contains different statistics of NBA players, for instance, players' position, age, assists, steals, etc.
 There are in total of 718 observations before any data cleanings.
 
-Since this dataset contains 3-point field goals, 2-point field goals, free throws and field goals, points can be calculated solely using these variables.
-My goal is to predict a the numebr of scores a NBA player scores based on his performance, which means that these features are too closely related to the target variable (i.e.
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-top5players
+#| tbl-cap: Top 5 NBA Players Based on Select Predictors Highly Correlated with Points Scored, Season 2023-2024
+
+raw_data <- read_csv("../data/raw_data/raw_data.csv")
+
+raw_data |> 
+  drop_na() |>
+  clean_names() |>
+  head(5) |> 
+  select(player, x3p, x2p, ft, pts) |>
+  rename(
+    Player = player,
+    "3-point goals" = x3p,
+    "2-point goals" = x2p,
+    "free throws" = ft,
+    "points" = pts
+  ) |>
+  kable()
+```
+
+This dataset contains player statistics including 3-point goals, 2-point goals, field goals, and free throws, points can be calculated using these variables.
+My goal is to predict the number of points a NBA player scores based on his performance, which means that these features are too closely related to the target variable (i.e.
 points).
+@tbl-top5players shows the table with closely related variables, we can calculate the points a player scored directly using these variables.
+For instance, player Bam Adebayo had 10 3-points goals, 470 2-point goals and 276 free throws during the 2023-2024 season, so his total points gained is $10 \cdot 3 + 470 \cdot 2 + 1 \cdot 276 = 1246$, which is exactly the points he scored.
 Thus, those variables are removed from the dataset using 'tidyverse' package in R @citeTidyverse.
-Moreover, there are 12 different positions in the dataset, I categorized them into a broader classification which has only three categories - Guards (G): SG, PG, SG-PG, PG-SG, Forwards (F): SF, PF, PF-SF, SF-PF, SF-SG, Centers (C): C, PF-C, C-PF.
+Moreover, there are 12 different positions in the dataset, I categorized them into a broader classification which has only three categories - Guards (G): SG, PG, SG-PG, PG-SG, Forwards (F): SF, PF, PF-SF, SF-PF, SF-SG, and Centers (C): C, PF-C, C-PF.
 This is done so that I don't need to create overwhelming dummy variables when generating the model and this process of binning could simplify the model and potentially strengthen the signal by reducing noise and multicollinearity.
+Additionally, redundant variables are also removed like
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-top10players
+#| tbl-cap: Top 10 NBA Players with selectely statistics, Season 2023-2024
+
+analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
+
+analysis_data |>
+  head(10) |>
+  kable()
+```
+
+After removing the closely related features, the dataset contains the following variables as @tbl-top10players shows:
+1. player: the player name, which includes the players that played in the 2023-2024 season.
+2. age: the age of each player.
+3. g: games, how many games a player played in this season.
+4. gs: game started, how many games a player has been in the starting lineup for their team at the beginning of the game.
+5. mp: minutes played, the total time of a player played in this season.
+6. fg_percent: field goal percentage, this statistic represents the percentage of field goals (both 2-pointers and 3-pointers) made by a player out of the total number of field goal attempts.
+7. x3p_percent: 3-point goal percentage, this statistic represents the percentage of 3-point field goals made by a player out of the total number of 3-point field goal attempts.
+8. x2p_percent: 2-point goal percentage, this statistic represents the percentage of 2-point field goals made by a player out of the total number of 2-point field goal attempts.
+9. e_fg_percent: effective field goal percentage, this statistic adjusts for the fact that a 3-point field goal is worth more than a 2-point field goal.
+10. ft_percent: free throw percentage, this statistic represents the percentage of free throws made by a player out of the total number of free throw attempts.
+11. orb: offensive rebounds, this statistic represents the number of rebounds grabbed by a player on the offensive end of the court.
+12. drb: defensive rebounds, this statistic represents the number of rebounds grabbed by a player on the defensive end of the court.
+13. trb: total rebounds, this statistic represents the total number of rebounds grabbed by a player (both offensive and defensive rebounds).
 
@@
 
-### Model justification
+## Model justification
 

### 2024-04-09 paper/paper.qmd
update

--- 2024-04-02 paper/paper.qmd
@@
 number-sections: true
+toc: true
 bibliography: references.bib
@@
 library(ggplot2)
+library(knitr)
+library(janitor)
+library(kableExtra)
 ```
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+You can and should cross-reference sections and sub-sections. We use @citeR and @citeRohan.
 
@@
 The dataset is directly downloaded from Basketball Reference using the given instruction.
-The dataset contains different statistics of NBA players, for instance players' position, age, assists, steals, etc.
+The dataset contains different statistics of NBA players, for instance, players' position, age, assists, steals, etc.
 There are in total of 718 observations before any data cleanings.
 
-Since this dataset contains 3-point field goals, 2-point field goals, free throws and field goals, points can be calculated solely using these variables.
-My goal is to predict a the numebr of scores a NBA player scores based on his performance, which means that these features are too closely related to the target variable (i.e.
+Raw data is displayed in @sec-raw-data, which is split into four different tables @tbl-tbl1, @tbl-tbl2, @tbl-tbl3, @tbl-tbl4.
+There are 30 variables in the dataset and each is introduced and explained in @sec-raw-data.
+
+## Data processing and interested predictors
+
+This dataset contains player statistics including 3-point goals, 2-point goals, field goals, and free throws, points can be calculated using these variables.
+My goal is to predict the number of points a NBA player scores based on his performance, which means that these features are too closely related to the target variable (i.e.
 points).
-Thus, those variables are removed from the dataset using 'tidyverse' package in R @citeTidyverse.
-Moreover, there are 12 different positions in the dataset, I categorized them into a broader classification which has only three categories - Guards (G): SG, PG, SG-PG, PG-SG, Forwards (F): SF, PF, PF-SF, SF-PF, SF-SG, Centers (C): C, PF-C, C-PF.
-This is done so that I don't need to create overwhelming dummy variables when generating the model and this process of binning could simplify the model and potentially strengthen the signal by reducing noise and multicollinearity.
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-top5players
+#| tbl-cap: Top 5 NBA Players Based on Select Predictors Highly Correlated with Points Scored, Season 2023-2024
+
+raw_data <- read_csv("../data/raw_data/raw_data.csv")
+
+raw_data |> 
+  drop_na() |>
+  clean_names() |>
+  head(5) |> 
+  select(player, x3p, x2p, ft, pts) |>
+  rename(
+    Player = player,
+    "3-point goals" = x3p,
+    "2-point goals" = x2p,
+    "free throws" = ft,
+    "points" = pts
+  ) |>
+  kable()
+```
+
+In @tbl-top5players, we see variables that are directly linked to a player's total points scored.
+For example, considering Bam Adebayo's performance in the 2023-2024 season: he scored 10 three-pointers, made 470 two-pointers, and successfully shot 276 free throws, totaling $10 \times 3 + 470 \times 2 + 276 \times 1 = 1246$ points, which exactly matches his recorded total points.
+To streamline the dataset for analysis, I utilized the 'tidyverse' package in R to remove these variables, which are 'fg', 'fga', 'x3p', 'x3pa', 'x2p', 'x2pa', 'ft', and 'fta'.
+
+The dataset initially detailed players across 12 unique positions, which was quite detailed for modeling purposes.
+To simplify, I grouped these positions into three main categories: Guards (G): SG, PG, SG-PG, PG-SG, Forwards (F): SF, PF, PF-SF, and Centers (C): C, PF-C, C-PF.
+This grouping was intended to make the model clearer and to potentially improve its predictive power by reducing unnecessary complexity and avoiding overlap in variables.
+
+Additionally, I re-evaluated the necessity of certain variables such as 'trb' (total rebounds), 'player', and 'tm' (team).
+'trb', being the sum of 'orb' (offensive rebounds) and 'drb' (defensive rebounds), didn't provide additional insight and was thus omitted.
+The 'player' variable was removed in favor of 'rk' (rank), which sufficed for identifying players without duplicating information.
+Lastly, the 'tm' variable was excluded as the analysis didn't focus on team-specific performance, making the team data unnecessary for this study.
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-top10players
+#| tbl-cap: Top 10 NBA Players with selectely statistics, Season 2023-2024
+
+analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
+
+analysis_data |>
+  head(10) |>
+  rename(
+    'fg%' = fg_percent,
+    'x3p%' = x3p_percent,
+    'x2p%' = x2p_percent,
+    'efg%' = e_fg_percent,
+    'fg%' = fg_percent,
+    'ft%' = ft_percent,
+    'pos' = broad_position
+  ) |>
+  kable(digits = 2)
+```
+
+As @tbl-top10players shows, apart from 'rk' for identification purpose, these variables are what we interested in and will be used as predictors to predict the total points a player would potentially gain based on his performances like 3-point goal percentage, position and etc.
 
@@
 
+# Additional data details
+
+## Raw data {#sec-raw-data}
+raw data from basketball reference is split into four tables for a better view, which are displayed below
+
 ```{r}
-#| include: false
+#| echo: false
 #| warning: false
 #| message: false
+#| label: tbl-tbl1
+#| tbl-cap: Basic information and overall performance
 
-analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
+# read raw_data from the folder
+raw_data <- read_parquet("../data/raw_data/raw_data.parquet")
 
-predictors <- c("g", "gs", "mp", "fg_percent", "x3p_percent", 
-                "x2p_percent", "e_fg_percent", "ft_percent", 
-                "orb", "drb", "trb", "ast", "stl", "blk", "tov", "pf")
+# clean it
+raw_data <- raw_data |> 
+  drop_na() |>
+  clean_names()
 
-for (predictor in predictors) {
-  p <- ggplot(analysis_data, aes_string(x = predictor, y = "pts")) + 
-    geom_point(alpha = 0.5) + 
-    geom_smooth(method = "lm", se = FALSE, color = "blue") + 
-    ggtitle(paste("Points vs", predictor)) +
-    theme_minimal()
+# splitting the raw_data to 4 tables for a better view
+tbl1 <- raw_data |>
+  select(rk, player, pos, age, tm, g, gs, mp, pts) |>
+  head(5) |>
+  kable()
 
-  print(p)
-}
+tbl1
 ```
 
-# Additional data details
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-tbl2
+#| tbl-cap: Shooting efficiency
 
-# Model details {#sec-model-details}
+# read raw_data from the folder
+raw_data <- read_parquet("../data/raw_data/raw_data.parquet")
 
-## Posterior predictive check
+# clean it
+raw_data <- raw_data |> 
+  drop_na() |>
+  clean_names()
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+# splitting the raw_data to 4 tables for a better view
+tbl2 <- raw_data |>
+  select(player, fg, fga, fg_percent, x3p, x3pa, x3p_percent, x2p, x2pa, x2p_percent, e_fg_percent) |>
+  head(5) |>
+  kable()
 
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+tbl2
+```
 
-## Diagnostics
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-tbl3
+#| tbl-cap: Free throws and rebounds
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
+# read raw_data from the folder
+raw_data <- read_parquet("../data/raw_data/raw_data.parquet")
 
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+# clean it
+raw_data <- raw_data |> 
+  drop_na() |>
+  clean_names()
+
+# splitting the raw_data to 4 tables for a better view
+tbl3 <- raw_data |>
+  select(player, ft, fta, ft_percent, orb, drb, trb) |>
+  head(5) |>
+  kable()
+
+tbl3
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-tbl4
+#| tbl-cap: Playmaking and defence
+
+# read raw_data from the folder
+raw_data <- read_parquet("../data/raw_data/raw_data.parquet")
+
+# clean it
+raw_data <- raw_data |> 
+  drop_na() |>
+  clean_names()
+
+# splitting the raw_data to 4 tables for a better view
+tbl4 <- raw_data |>
+  select(player, ast, stl, blk, tov, pf) |>
+  head(5) |>
+  kable()
+
+tbl4
+```
+
+1. rk: rank - this doesn't represent the ranking of players based on some criterion, but purely for numbering purpose
+2. player: player - the name of the basketball player.
+
+3. pos: position - the playing position of the player.
+
+4. age: the age of each player.
+
+5. tm: team - the abbreviation of the NBA team the player belongs to.
+
+6. g: games - how many games a player played in this season.
+
+7. gs: game started - how many games a player has been in the starting lineup for their team at the beginning of the game.
+
+8. mp: minutes played - the total time of a player played in this season.
+
+9. fg: field goals - the total number of field goals (baskets) the player has made.
+
+10. fga: field goal attempts - the total number of field goal shots the player has attempted.
+
+11. fg_percent: field goal percentage - this statistic represents the percentage of field goals (both 2-pointers and 3-pointers) made by a player out of the total number of field goal attempts.
+
+12. x3p: 3-point field goals - the total number of 3-point field goals the player has made.
+
+13. x3pa: 3-point field goal attempts - the total number of 3-point shots the player has attempted.
+
+14. x3p_percent: 3-point goal percentage - this statistic represents the percentage of 3-point field goals made by a player out of the total number of 3-point field goal attempts.
+
+15. x2p: 2-point field goals - the total number of 2-point field goals the player has made.
+
+16. x2pa: 2-point field goal attempts - the total number of 2-point shots the player has attempted.
+
+17. x2p_percent: 2-point goal percentage - this statistic represents the percentage of 2-point field goals made by a player out of the total number of 2-point field goal attempts.
+
+18. e_fg_percent: effective field goal percentage - this statistic adjusts for the fact that a 3-point field goal is worth more than a 2-point field goal.
+
+19. ft: free throws - the total number of free throws the player has made.
+
+20: fta: free throw attempts - the total number of free throw shots the player has attempted.
+
+21. ft_percent: free throw percentage - this statistic represents the percentage of free throws made by a player out of the total number of free throw attempts.
+
+22. orb: offensive rebounds - this statistic represents the number of rebounds grabbed by a player on the offensive end of the court.
+
+23. drb: defensive rebounds - this statistic represents the number of rebounds grabbed by a player on the defensive end of the court.
+
+24. trb: total rebounds - this statistic represents the total number of rebounds grabbed by a player (both offensive and defensive rebounds).
+
+25. ast: assists - the total number of assists the player has made, indicating the number of times a player's pass led directly to a basket by a teammate.
+
+26. stl: steals - the total number of times the player has taken the ball away from an opponent, leading to a change in possession.
+
+27. blk: blocks - the total number of times the player has deflected an opponent's filed goal attempt, preventing the ball from going into the basket.
+
+28. tov: turnovers - the total number of times the player has lost possession of the ball to the opposing team.
+
+29. pf: personal fouls - the total number of personal fouls the player has committed.
+
+30: pts: points - the total number of points the player has scored.
 

### 2024-04-09 scripts/01-download_data.R

@@ -5,7 +5,18 @@
 # Contact: Yihang.cai.nz@gmail.com
 # Any other information needed? Some of the codes are adapted from Telling Stories with Data by Rohan Alexander
 
+#### Workspace setup ####
+library(tidyverse)
+library(arrow)
 
 # raw data is downloaded directed from basketball reference on this website: https://www.basketball-reference.com/leagues/NBA_2024_totals.html
+raw_data <- read_csv("data/raw_data/raw_data.csv")
+
+# convert the dataset to UTF-8
+raw_data <- raw_data |>
+  mutate(across(where(is.character), ~iconv(., from = "Windows-1252", to = "UTF-8")))
+
+
+write_parquet(raw_data, "data/raw_data/raw_data.parquet")
 
          

### 2024-04-09 scripts/02-data_cleaning.R

@@ -25,7 +25,7 @@ cleaned_data <-
     pos %in% c("C", "PF-C", "C-PF") ~ "C"
     )
   ) |>
-  select(-rk, -tm, -pos, -fg, -fga, -x3p, -x3pa, -x2p, -x2pa, -ft, -fta) |>
+  select(-player, -tm, -pos, -fg, -fga, -x3p, -x3pa, -x2p, -x2pa, -ft, -fta, -trb) |>
   tidyr::drop_na()
 
 

### 2024-04-09 paper/paper.qmd
model section completed

@@
 
-The dataset is obtained from Basketball Reference @citeDataset, which contains NBA players' statistics in the 2023 to 2024 season.
-The dataset is directly downloaded from Basketball Reference using the given instruction.
-The dataset contains different statistics of NBA players, for instance, players' position, age, assists, steals, etc.
-There are in total of 718 observations before any data cleanings.
+The dataset for this analysis was acquired from Basketball Reference @citeDataset and includes a wide range of NBA player statistics for the 2023-2024 season.
+The process of downloading this dataset involved converting the website's data table into a CSV format, then transferring this data into Excel.
+In Excel, I employed the 'Text to Columns' feature to separate the statistics using commas, thereby preparing the dataset for analysis.
+This dataset comprises a variety of player statistics, such as position, age, assists, and steals, with a total of 718 observations before any data cleaning.
 
-Raw data is displayed in @sec-raw-data, which is split into four different tables @tbl-tbl1, @tbl-tbl2, @tbl-tbl3, @tbl-tbl4.
-There are 30 variables in the dataset and each is introduced and explained in @sec-raw-data.
+The analysis was conducted in the R statistical programming environment @citeR, utilizing a selection of packages for different tasks.
+Data cleaning was performed using the 'Tidyverse' @citeTidyverse and 'Janitor' @citeJanitor packages, while 'Dplyr' @citeDplyr was used for data manipulation.
+The 'Knitr' @citeKnitr and 'Ggplot2' @citeGgplot2 packages were employed for data visualization, including the creation of tables and figures.
+Predictive modeling and model visualization were facilitated by the 'Tidymodels' @citeTidymodels and 'Modelsummary' @citeModelsummary packages, respectively.
+
+The raw data is presented in @sec-raw-data, divided into four separate tables (@tbl-tbl1, @tbl-tbl2, @tbl-tbl3, @tbl-tbl4).
+The dataset contains 30 variables, each thoroughly introduced and explained in @sec-raw-data, offering a detailed view of the data that forms the basis of this study.
 
@@
 
-This dataset contains player statistics including 3-point goals, 2-point goals, field goals, and free throws, points can be calculated using these variables.
-My goal is to predict the number of points a NBA player scores based on his performance, which means that these features are too closely related to the target variable (i.e.
-points).
+This dataset comprises statistics for players, such as 3-point goals, 2-point goals, field goals, and free throws, from which points can be derived.
+My objective is to forecast an NBA player's total points based on their performance metrics, indicating a direct correlation between these features and the target variable, namely, points.
 
@@
 
+# read in the raw_data
 raw_data <- read_csv("../data/raw_data/raw_data.csv")
 
+# cleans the raw_data for universal variable name and then make a table using kable function
 raw_data |> 
@@
 
+# read in the analysis_data
 analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
 
+# show the first 10 players, and then rename some of the variables for a better view, and set the digits to two d.p. for a better view
 analysis_data |>
@@
 
-As @tbl-top10players shows, apart from 'rk' for identification purpose, these variables are what we interested in and will be used as predictors to predict the total points a player would potentially gain based on his performances like 3-point goal percentage, position and etc.
+As illustrated in @tbl-top10players, aside from 'rk' serving as an identifier, the selected variables are central to our analysis.
+These will act as predictors for estimating a player's total points, considering factors such as 3-point goal percentage, position, among others.
+Furthermore, for enhanced clarity, the values in the tables have been formatted to display two decimal places.
 
@@
 
-Lasso regression is a variation of linear regressions that can actually perform feature selection by setting the coefficients of less important features to zero.
-It has an additional parameter $\lambda$, which is the regularization parameter that controls the strength of the penalty.
-The penalty term shrinks some of the coefficients (i.e.
-the less important variables) toward zero.
-As $\lambda$ increases, more coefficients are set to zero, leading to a simpler model.
-The value of $\lambda$ is determined through cross-validation.
+This model pursues two main aims.
+The initial goal is to predict the total points an NBA player might score based on various performance indicators such as position and shooting efficiency.
+The second goal is to identify which predictors most significantly affect a player's scoring ability.
+For example, it is assumed that more playtime within a season could lead to a higher score.
+
+To meet these objectives, the lasso regression model is chosen for its unique features: First, with 19 predictors left after processing the data, the lasso regression can reduce the influence of less important predictors by setting their coefficients to zero.
+Second, it clearly indicates which predictors have a greater impact on the scoring outcome, helping to understand what factors are most important in determining a player's points.
+
+Lasso regression, a variant of linear regression models, is notable for its ability to select features by reducing the coefficients of less critical features to zero.
+This model introduces a regularization parameter, $\lambda$, which determines the strength of the penalty.
+This penalty minimizes some coefficients, especially those for less important variables, towards zero.
+As $\lambda$ increases, more coefficients are reduced to zero, leading to a simpler model.
+The optimal value for $\lambda$ is determined through cross-validation, ensuring the model is effectively tuned for the predictive tasks.
 
@@
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+We anticipate a positive correlation between the points scored and several factors: age, minutes played, number of games played, games started, shooting efficiency (encompassed by 2-point goal percentage, 3-point goal percentage, field goal percentage, and free throw percentage), rebounds (both offensive and defensive), and assists.
+The logic is straightforward: the higher these variables, the greater the likelihood of scoring more points.
+Additionally, a player's position could influence their scoring, as different positions in basketball have distinct objectives; for instance, center(C) may prioritize defense over scoring.
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+## Model performance
 
 ```{r}
-# #| include: false
-# #| warning: false
-# #| message: false
-# 
-# first_lasso_model <- readRDS("../models/first_lasso_model.rds")
-# second_lasso_model <- readRDS("../models/second_lasso_model.rds")
-# predictions <- predict(second_lasso_model, test_data)
-# results <- bind_cols(test_data, predictions)
-# # Calculate RMSE
-# rmse_results_2 <- rmse(results, truth = pts, estimate = .pred)
-# rsq_results_2 <- rsq(results, truth = pts, estimate = .pred)
+#| include: false
+#| warning: false
+#| message: false
+#| label: tbl-rmse
+#| tbl-cap: Comparison of Prediction Accuracy for NBA Player Points: Initial vs. Tuned Lasso Models
+
+first_lasso_model <- readRDS("../models/first_lasso_model.rds")
+second_lasso_model <- readRDS("../models/second_lasso_model.rds")
+
+analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
+
+# data splitting
+set.seed(321) # For reproducibility
+data_split <- initial_split(analysis_data, prop = 0.8)
+test_data <- testing(data_split)
+
+# predictions for the first model
+predictions_1 <- predict(first_lasso_model, test_data)
+results_1 <- bind_cols(test_data, predictions_1)
+# Calculate RMSE for the first model
+rmse_results_1 <- rmse(results_1, truth = pts, estimate = .pred)
+
+analysis_data_1 <- read_parquet("../data/analysis_data/analysis_data.parquet")
+analysis_data_1$age_squared = analysis_data_1$age^2
+analysis_data_1$pts_per_min = analysis_data_1$pts / analysis_data_1$mp
+analysis_data_1$ast_per_min = analysis_data_1$ast / analysis_data_1$mp
+
+set.seed(123) # For reproducibility
+data_split <- initial_split(analysis_data_1, prop = 0.8)
+train_data <- training(data_split)
+test_data <- testing(data_split)
+
+# predictions for the second model
+predictions_2 <- predict(second_lasso_model, test_data)
+results_2 <- bind_cols(test_data, predictions_2)
+# Calculate RMSE for the first model
+rmse_results_2 <- rmse(results_2, truth = pts, estimate = .pred)
+
+# Create a data frame with the RMSE values
+rmse_comparison <- data.frame(
+  "Initial Model" = rmse_results_1$.estimate,
+  "Tuned Model" = rmse_results_2$.estimate
+)
+
+# Set row names to 'RMSE' for clarity
+rownames(rmse_comparison) <- "RMSE"
+
+# Display the table
+kable(rmse_comparison)
 ```
@@
 # Results
-
-Our results are summarized in @tbl-modelresults.
 

### 2024-04-09 scripts/04-model.R

@@ -24,7 +24,7 @@ test_data <- testing(data_split)
 
 # preprocesssing with recipe
 data_recipe <- recipe(pts ~ ., data = train_data) |>
-  step_rm(player) |>
+  step_rm(rk) |>
   step_normalize(all_numeric_predictors()) |>
   step_dummy(broad_position)
 
@@ -63,7 +63,6 @@ final_lasso_workflow <- finalize_workflow(
 first_lasso_model <- fit(final_lasso_workflow, data = train_data)
 
 # Second model with feature engineering
-
 analysis_data_1 <- read_parquet("data/analysis_data/analysis_data.parquet")
 analysis_data_1$age_squared = analysis_data_1$age^2
 analysis_data_1$pts_per_min = analysis_data_1$pts / analysis_data_1$mp
@@ -78,7 +77,7 @@ test_data <- testing(data_split)
 
 # preprocesssing with recipe
 data_recipe <- recipe(pts ~ ., data = train_data) |>
-  step_rm(player) |>
+  step_rm(rk) |>
   step_normalize(all_numeric_predictors()) |>
   step_dummy(broad_position)
 

### 2024-04-09 paper/paper.qmd
update

@@
 ```{r}
-#| include: false
+#| echo: false
 #| warning: false
 #| message: false
-#| label: tbl-rmse
-#| tbl-cap: Comparison of Prediction Accuracy for NBA Player Points: Initial vs. Tuned Lasso Models
+#| label: tbl-model1
+#| tbl-cap: First lasso regression model top 10 predictions
 
 first_lasso_model <- readRDS("../models/first_lasso_model.rds")
-second_lasso_model <- readRDS("../models/second_lasso_model.rds")
 
@@
 # predictions for the first model
-predictions_1 <- predict(first_lasso_model, test_data)
-results_1 <- bind_cols(test_data, predictions_1)
-# Calculate RMSE for the first model
-rmse_results_1 <- rmse(results_1, truth = pts, estimate = .pred)
+predictions <- predict(first_lasso_model, test_data)
+results <- bind_cols(test_data, predictions)
 
-analysis_data_1 <- read_parquet("../data/analysis_data/analysis_data.parquet")
-analysis_data_1$age_squared = analysis_data_1$age^2
-analysis_data_1$pts_per_min = analysis_data_1$pts / analysis_data_1$mp
-analysis_data_1$ast_per_min = analysis_data_1$ast / analysis_data_1$mp
+results |>
+  select(rk, pts, .pred) |>
+  head(10) |>
+  rename(
+    "Rank" = rk,
+    "Points" = pts,
+    "Prediction" = .pred
+    ) |>
+  kable(booktabs = TRUE, digits = 2)
+```
+@tbl-model1 displays the predictions made by the initial lasso regression model for the number of points scored by NBA players, numbered by 'Rank' identifier.
+For each 'Rank', there are two columns: 'Points', which represents the actual points scored, and 'Prediction', which shows the predicted points scored by the model.
+It's noticeable that there's a variance between the actual points and the predicted values.
+The model does not seem to accurately predict the points: In some cases, such as Rank 3, the model overestimates the points, predicting 372.78 points against the actual 274.
+In other instances, like Rank 7, the prediction is quite close to the actual points scored (1128.62 predicted vs.
+1142 actual).
+There are also underestimations, as seen with Rank 30, where the model predicts 985.43 points while the actual points scored are 1036.
 
-set.seed(123) # For reproducibility
-data_split <- initial_split(analysis_data_1, prop = 0.8)
-train_data <- training(data_split)
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-model1rmsemae
+#| tbl-cap: First lasso regression model RMSE and MAE
+
+# read in the first lasso model
+first_lasso_model <- readRDS("../models/first_lasso_model.rds")
+
+analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
+
+# data splitting
+set.seed(321) # For reproducibility
+data_split <- initial_split(analysis_data, prop = 0.8)
 test_data <- testing(data_split)
 
-# predictions for the second model
-predictions_2 <- predict(second_lasso_model, test_data)
-results_2 <- bind_cols(test_data, predictions_2)
+# predictions for the first model
+predictions <- predict(first_lasso_model, test_data)
+results <- bind_cols(test_data, predictions)
 # Calculate RMSE for the first model
-rmse_results_2 <- rmse(results_2, truth = pts, estimate = .pred)
+rmse_results <- rmse(results, truth = pts, estimate = .pred)
+mae_results <- mae(results, pts, .pred)
 
-# Create a data frame with the RMSE values
-rmse_comparison <- data.frame(
-  "Initial Model" = rmse_results_1$.estimate,
-  "Tuned Model" = rmse_results_2$.estimate
+results <- data.frame(
+  "RMSE" = rmse_results$.estimate,
+  "MAE" = mae_results$.estimate
 )
 
-# Set row names to 'RMSE' for clarity
-rownames(rmse_comparison) <- "RMSE"
+rownames(results) <- "First lasso regression model"
 
-# Display the table
-kable(rmse_comparison)
+kable(results, booktabs = TRUE, digits = 2)
+```
+
+The table @tbl-model1rmsemae lists two error metrics for assessing the first lasso regression model.
+RMSE(Root Mean Squared Error), recorded at 111.79, captures the average error by squaring the difference between the model's predictions and the actual points, thereby giving more weight to larger discrepancies and making it particularly useful where such errors have greater consequences.
+MAE(Mean Absolute Error), noted as 80.6, represents the simple average of all prediction errors without emphasis on their size, making it a reliable metric when treating all errors uniformly is preferable.
+Utilizing both RMSE and MAE provides a dual perspective: RMSE highlights the impact of substantial errors, and MAE offers a clear measure of average error, assisting in a balanced evaluation of the model's performance.
+This approach to error analysis suggests that the model's predictions could be improved by reevaluating the included features, especially in areas where the model's accuracy is critical.
+
+### Feature engineering
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-model1variables
+#| tbl-cap: First lasso regression model important variables
+
 ```

### 2024-04-09 scripts/04-model.R

@@ -128,3 +128,27 @@ saveRDS(
 )
 
 
+
+library(tidymodels)
+library(broom)
+
+# Extract the fitted model object
+lasso_fit <- extract_fit_parsnip(first_lasso_model)
+
+# Use broom's tidy method on the extracted fit object
+coef_info <- tidy(lasso_fit)
+
+# Now, filter out the intercept and examine the coefficients
+important_predictors <- coef_info %>%
+  filter(term != "(Intercept)") %>%
+  arrange(desc(abs(estimate)))
+
+# View the important predictors
+print(important_predictors)
+
+
+
+
+
+
+

### 2024-04-10 paper/paper.qmd
update

@@
 The analysis was conducted in the R statistical programming environment @citeR, utilizing a selection of packages for different tasks.
-Data cleaning was performed using the 'Tidyverse' @citeTidyverse and 'Janitor' @citeJanitor packages, while 'Dplyr' @citeDplyr was used for data manipulation.
+Data cleaning was performed using the 'Tidyverse' @citeTidyverse and 'Janitor' @citeJanitor packages, while 'Dplyr' @citeDplyr and 'Broom' @citeBroom were used for data manipulation and data frame visualization.
 The 'Knitr' @citeKnitr and 'Ggplot2' @citeGgplot2 packages were employed for data visualization, including the creation of tables and figures.
@@
 
+first_lasso_model <- readRDS("../models/first_lasso_model.rds")
+
+# Extract the fitted model object
+lasso_fit <- extract_fit_parsnip(first_lasso_model)
+
+# Use broom's tidy method on the extracted fit object
+coef_info <- tidy(lasso_fit)
+
+# Now, filter out the intercept and examine the coefficients
+important_predictors <- coef_info |>
+  filter(term != "(Intercept)") |>
+  arrange(desc(abs(estimate)))
+
+# View the important predictors
+important_predictors |>
+  select(-penalty) |>
+  rename(
+    "Predictors" = term,
+    "Coefficients" = estimate
+  ) |>
+  kable(booktabs = TRUE, digits = 2)
 ```

### 2024-04-10 scripts/04-model.R

@@ -129,22 +129,6 @@ saveRDS(
 
 
 
-library(tidymodels)
-library(broom)
-
-# Extract the fitted model object
-lasso_fit <- extract_fit_parsnip(first_lasso_model)
-
-# Use broom's tidy method on the extracted fit object
-coef_info <- tidy(lasso_fit)
-
-# Now, filter out the intercept and examine the coefficients
-important_predictors <- coef_info %>%
-  filter(term != "(Intercept)") %>%
-  arrange(desc(abs(estimate)))
-
-# View the important predictors
-print(important_predictors)
 
 
 

### 2024-04-12 paper/paper.qmd
feature engineering finished

@@
 #| label: tbl-model1variables
-#| tbl-cap: First lasso regression model important variables
+#| tbl-cap: Top 5 important variables of the first lasso regression model
 
@@
 important_predictors |>
+  head(5) |>
   select(-penalty) |>
@@
 ```
+
+@tbl-modelvariables lists the predictors with the highest magnitude coefficients from the lasso regression model, indicating their relative importance in predicting the outcome variable.
+The listed predictors are minutes played ('mp'), turnovers ('tov'), defensive rebounds ('drb'), offensive rebounds ('orb'), and personal fouls ('pf'), with their corresponding coefficients.
+
+Minutes Played (310.86): The positive coefficient suggests a strong positive relationship between the number of minutes a player plays and their point total.
+More minutes played typically allows more opportunities to score points.
+Turnovers (225.98): The positive coefficient here is interesting as turnovers are generally negative events.
+However, it might indicate that players who are more involved in the game (and therefore more likely to score points) also have more turnovers.
+Defensive Rebounds (73.62): A positive coefficient for defensive rebounds implies players who gather more rebounds on the defensive end also tend to score more points, possibly due to the additional possessions they secure for their team.
+Offensive Rebounds (-70.80): The negative coefficient for offensive rebounds suggests that players who get more offensive rebounds do not necessarily score more points.
+This could reflect a role specialization where some players focus on rebounding rather than scoring.
+Personal Fouls (-65.66): The negative coefficient indicates that players who commit more fouls tend to score fewer points.
+This could be because committing fouls may limit a player's time on the court due to foul trouble.
 

### 2024-04-12 scripts/04-model.R

@@ -64,9 +64,13 @@ first_lasso_model <- fit(final_lasso_workflow, data = train_data)
 
 # Second model with feature engineering
 analysis_data_1 <- read_parquet("data/analysis_data/analysis_data.parquet")
-analysis_data_1$age_squared = analysis_data_1$age^2
-analysis_data_1$pts_per_min = analysis_data_1$pts / analysis_data_1$mp
-analysis_data_1$ast_per_min = analysis_data_1$ast / analysis_data_1$mp
+analysis_data_1 <- analysis_data_1 |>
+  mutate(
+    points_per_min = pts / mp,
+    tov_per_game = tov /g,
+    pf_per_game = pf / g
+  )
+
 
 ### Model data ####
 
@@ -115,6 +119,18 @@ final_lasso_workflow <- finalize_workflow(
 # fit the model
 second_lasso_model <- fit(final_lasso_workflow, data = train_data)
 
+predictions <- predict(second_lasso_model, test_data)
+results <- bind_cols(test_data, predictions)
+# Calculate RMSE for the first model
+rmse_results <- rmse(results, truth = pts, estimate = .pred)
+mae_results <- mae(results, pts, .pred)
+
+results <- data.frame(
+  "RMSE" = rmse_results$.estimate,
+  "MAE" = mae_results$.estimate
+)
+
+results
 
 #### Save model ####
 saveRDS(

### 2024-04-12 paper/paper.qmd
update

--- 2024-04-10 paper/paper.qmd
@@
 library(janitor)
-library(kableExtra)
 ```
@@
 #| label: tbl-model1rmsemae
-#| tbl-cap: First lasso regression model RMSE and MAE
+#| tbl-cap: RMSE and MAE of first lasso regression model
 
@@
 
+@tbl-model1variables lists the predictors with the highest magnitude coefficients from the lasso regression model, indicating their relative importance in predicting the outcome variable.
+The listed predictors are minutes played ('mp'), turnovers ('tov'), defensive rebounds ('drb'), offensive rebounds ('orb'), and personal fouls ('pf'), with their corresponding coefficients.
+
+The coefficient for 'mp' is positive (310.86), highlighting a direct relationship with point totals — more minutes played usually provides more opportunities for scoring.
+In contrast, 'orb' has a negative coefficient (-70.80), hinting at players with high offensive rebounds not necessarily correlating with higher points, perhaps indicating a focus on rebounding over scoring.
+'Tov' carries a positive coefficient (225.98), which might seem counterintuitive given turnovers are adverse events; yet, it could reflect that players who handle the ball frequently might incur more turnovers and also have more scoring chances.
+A positive coefficient for 'drb' (73.62) suggests a link between securing defensive rebounds and higher point scores, likely due to the additional possessions gained.
+'Pf' shows a negative coefficient (-65.66), indicating that fouling frequently could decrease a player's scoring by reducing playing time due to foul trouble.
+
+To refine the model, feature engineering introduced the 'pts_per_min' predictor, combining points with minutes played to assess scoring efficiency.
+This reflects how well players score relative to their time on the court.
+'tov_per_game' adjusts turnovers for the number of games, enabling a fairer comparison across players, and 'pf_per_game' computes the average fouls per game, a significant aspect in evaluating defensive conduct and the potential impact on game participation and point contribution.
+These engineered features aim to provide a clearer understanding of each player's performance, leading to an improved model with lower error metrics.
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-model2rmsemae
+#| tbl-cap: RMSE and MAE of finalized lasso regression model
+
+# read in the first lasso model
+second_lasso_model <- readRDS("../models/second_lasso_model.rds")
+
+analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
+
+analysis_data <- analysis_data |>
+  mutate(
+    pts_per_min = pts / mp,
+    tov_per_game = tov /g,
+    pf_per_game = pf / g
+  )
+
+# data splitting
+set.seed(123) # For reproducibility
+data_split <- initial_split(analysis_data, prop = 0.8)
+test_data <- testing(data_split)
+
+# predictions for the first model
+predictions <- predict(second_lasso_model, test_data)
+results <- bind_cols(test_data, predictions)
+# Calculate RMSE for the first model
+rmse_results <- rmse(results, truth = pts, estimate = .pred)
+mae_results <- mae(results, pts, .pred)
+
+results <- data.frame(
+  "RMSE" = rmse_results$.estimate,
+  "MAE" = mae_results$.estimate
+)
+
+rownames(results) <- "Second lasso regression model"
+
+kable(results, booktabs = TRUE, digits = 2)
+```
+
+As shown in @tbl-model2rmsemae, the second lasso regression model, enhanced with engineered features, has demonstrated significant improvement.
+The RMSE has decreased from 111.7 to 94.68, and the MAE has dropped from 80.6 to 66.14.
+This reduction in both metrics indicates that the model now predicts the number of points an NBA player could score based on their performances with greater precision.
+
 # Results
+
+For the Results section of your paper based on the rubric and the information provided, you should aim to present a clear, concise, and informative summary of the findings from your lasso regression model analysis.
+Here are structured suggestions on how to frame your Results section effectively:
+
+Results Section Framework:
+Overview of Model Performance:
+
+Start by summarizing the overall performance of your final lasso regression model.
+Mention the improvement in the error metrics (RMSE and MAE) after the feature engineering, highlighting how these changes have enhanced the model's predictive accuracy.
+Example: "The refinement of the lasso regression model through targeted feature engineering significantly enhanced its predictive accuracy.
+The RMSE decreased from 111.7 to 94.68, and the MAE improved from 80.6 to 66.14, indicating a more precise model in predicting NBA players' points based on performance metrics."
+Detailed Results from Predictive Analysis:
+
+Present specific results, such as how well the model predicted points for different players or player types.
+If applicable, include comparisons to show the model's performance before and after feature engineering.
+Example: "Detailed analysis reveals that the model, after enhancements, provides a more accurate prediction across a range of player performance indicators, such as minutes played and turnovers.
+The incorporation of efficiency metrics like points per minute played ('pts_per_min') significantly contributed to this improvement."
+Statistical Significance and Confidence:
+
+Discuss any statistical tests or confidence intervals used to ascertain the reliability of the model's predictions.
+This could involve discussing p-values for coefficients or the confidence in the model's error metrics.
+Example: "Statistical tests affirm the significance of newly integrated predictors, with p-values well below the conventional threshold of 0.05, thereby reinforcing confidence in the model's enhancements."
+Visualization of Results:
+
+If visual aids (such as plots or charts) were used to illustrate the model's performance or to compare the predicted vs. actual points, briefly describe these visualizations and their outcomes.
+Example: "Graphical representations, included in fig-predictions, show a tight correlation between the model's predictions and the actual points scored, especially after the feature engineering phase."
+Summary of Key Findings:
+
+Conclude with a summary of the most critical findings from the model's application, emphasizing how these insights support the model's utility in a real-world context.
+Example: "The results confirm that the enhanced lasso regression model not only predicts the point outcomes with greater accuracy but also highlights the importance of game participation metrics such as minutes played and turnovers in determining a player's scoring potential."
+Additional Tips:
+Clarity and Precision: Keep the language clear and precise, focusing on what the reader needs to understand about the model's performance and the significance of the results.
+Referencing Tables/Graphs: Ensure to reference any tables or graphs from your analysis that support your results.
+This helps in validating your claims and provides a visual or tabular anchor for your findings.
+Contextualization: Briefly relate your findings to the broader context of sports analytics, perhaps suggesting how these insights could influence team strategy or player evaluation.
+This structure and approach should align with the academic standards outlined in the rubric, ensuring that your Results section is not only informative but also well-integrated with the rest of your paper.
 

### 2024-04-12 scripts/04-model.R

@@ -66,11 +66,10 @@ first_lasso_model <- fit(final_lasso_workflow, data = train_data)
 analysis_data_1 <- read_parquet("data/analysis_data/analysis_data.parquet")
 analysis_data_1 <- analysis_data_1 |>
   mutate(
-    points_per_min = pts / mp,
+    pts_per_min = pts / mp,
     tov_per_game = tov /g,
     pf_per_game = pf / g
   )
-
 
 ### Model data ####
 
@@ -118,19 +117,6 @@ final_lasso_workflow <- finalize_workflow(
 
 # fit the model
 second_lasso_model <- fit(final_lasso_workflow, data = train_data)
-
-predictions <- predict(second_lasso_model, test_data)
-results <- bind_cols(test_data, predictions)
-# Calculate RMSE for the first model
-rmse_results <- rmse(results, truth = pts, estimate = .pred)
-mae_results <- mae(results, pts, .pred)
-
-results <- data.frame(
-  "RMSE" = rmse_results$.estimate,
-  "MAE" = mae_results$.estimate
-)
-
-results
 
 #### Save model ####
 saveRDS(

### 2024-04-12 paper/paper.qmd
update

@@
 
-For the Results section of your paper based on the rubric and the information provided, you should aim to present a clear, concise, and informative summary of the findings from your lasso regression model analysis.
-Here are structured suggestions on how to frame your Results section effectively:
-
 Results Section Framework:
@@
 
-Start by summarizing the overall performance of your final lasso regression model.
+Start by summarizing the overall performance of the final lasso regression model.
 Mention the improvement in the error metrics (RMSE and MAE) after the feature engineering, highlighting how these changes have enhanced the model's predictive accuracy.

### 2024-04-17 paper/paper.qmd
update

@@
 The 'Knitr' @citeKnitr and 'Ggplot2' @citeGgplot2 packages were employed for data visualization, including the creation of tables and figures.
-Predictive modeling and model visualization were facilitated by the 'Tidymodels' @citeTidymodels and 'Modelsummary' @citeModelsummary packages, respectively.
+Predictive modeling was done by the 'Tidymodels' @citeTidymodels.
 
@@
 The 'player' variable was removed in favor of 'rk' (rank), which sufficed for identifying players without duplicating information.
+Additionally, there are duplicated ranks ('rk') since some players may be transferred to a different team during the season, so the identifier is not unique.
 Lastly, the 'tm' variable was excluded as the analysis didn't focus on team-specific performance, making the team data unnecessary for this study.
@@
 
-# Model
+# Model {#sec-Model}
 
-This model pursues two main aims.
+This model pursues two main goals.
 The initial goal is to predict the total points an NBA player might score based on various performance indicators such as position and shooting efficiency.
@@
 # data splitting
-set.seed(321) # For reproducibility
+set.seed(123) # For reproducibility
 data_split <- initial_split(analysis_data, prop = 0.8)
@@
 
-The table @tbl-model1rmsemae lists two error metrics for assessing the first lasso regression model.
+@tbl-model1rmsemae lists two error metrics for assessing the first lasso regression model.
 RMSE(Root Mean Squared Error), recorded at 111.79, captures the average error by squaring the difference between the model's predictions and the actual points, thereby giving more weight to larger discrepancies and making it particularly useful where such errors have greater consequences.
@@
 'tov_per_game' adjusts turnovers for the number of games, enabling a fairer comparison across players, and 'pf_per_game' computes the average fouls per game, a significant aspect in evaluating defensive conduct and the potential impact on game participation and point contribution.
+
+To guarantee that the final lasso regression model's predictions stay within realistic limits, we've implemented a simple yet deliberate adjustment: all negative predicted values are set to zero.
+While this approach may appear overly simplistic, it is employed for considered reasons that will be elaborated upon in the Discussion section (@sec-Discussion).
+
 These engineered features aim to provide a clearer understanding of each player's performance, leading to an improved model with lower error metrics.
+
+# Results
+
+## Model overview
 
@@
 predictions <- predict(second_lasso_model, test_data)
+
+predictions[predictions < 0] <- 0
+
 results <- bind_cols(test_data, predictions)
@@
 As shown in @tbl-model2rmsemae, the second lasso regression model, enhanced with engineered features, has demonstrated significant improvement.
-The RMSE has decreased from 111.7 to 94.68, and the MAE has dropped from 80.6 to 66.14.
+The RMSE has decreased from 110.17 to 91.6, and the MAE has dropped from 78.2 to 61.31.
 This reduction in both metrics indicates that the model now predicts the number of points an NBA player could score based on their performances with greater precision.
 
-# Results
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-model2
+#| tbl-cap: Final lasso regression model top 10 predictions
 
-Results Section Framework:
-Overview of Model Performance:
+second_lasso_model <- readRDS("../models/second_lasso_model.rds")
 
-Start by summarizing the overall performance of the final lasso regression model.
-Mention the improvement in the error metrics (RMSE and MAE) after the feature engineering, highlighting how these changes have enhanced the model's predictive accuracy.
-Example: "The refinement of the lasso regression model through targeted feature engineering significantly enhanced its predictive accuracy.
-The RMSE decreased from 111.7 to 94.68, and the MAE improved from 80.6 to 66.14, indicating a more precise model in predicting NBA players' points based on performance metrics."
-Detailed Results from Predictive Analysis:
+analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
 
-Present specific results, such as how well the model predicted points for different players or player types.
-If applicable, include comparisons to show the model's performance before and after feature engineering.
-Example: "Detailed analysis reveals that the model, after enhancements, provides a more accurate prediction across a range of player performance indicators, such as minutes played and turnovers.
-The incorporation of efficiency metrics like points per minute played ('pts_per_min') significantly contributed to this improvement."
-Statistical Significance and Confidence:
+analysis_data <- analysis_data |>
+  mutate(
+    pts_per_min = pts / mp,
+    tov_per_game = tov /g,
+    pf_per_game = pf / g
+  )
 
-Discuss any statistical tests or confidence intervals used to ascertain the reliability of the model's predictions.
-This could involve discussing p-values for coefficients or the confidence in the model's error metrics.
-Example: "Statistical tests affirm the significance of newly integrated predictors, with p-values well below the conventional threshold of 0.05, thereby reinforcing confidence in the model's enhancements."
-Visualization of Results:
+# data splitting
+set.seed(123) # For reproducibility
+data_split <- initial_split(analysis_data, prop = 0.8)
+test_data <- testing(data_split)
 
-If visual aids (such as plots or charts) were used to illustrate the model's performance or to compare the predicted vs. actual points, briefly describe these visualizations and their outcomes.
-Example: "Graphical representations, included in fig-predictions, show a tight correlation between the model's predictions and the actual points scored, especially after the feature engineering phase."
-Summary of Key Findings:
+# predictions for the first model
+predictions <- predict(second_lasso_model, test_data)
 
-Conclude with a summary of the most critical findings from the model's application, emphasizing how these insights support the model's utility in a real-world context.
-Example: "The results confirm that the enhanced lasso regression model not only predicts the point outcomes with greater accuracy but also highlights the importance of game participation metrics such as minutes played and turnovers in determining a player's scoring potential."
-Additional Tips:
-Clarity and Precision: Keep the language clear and precise, focusing on what the reader needs to understand about the model's performance and the significance of the results.
-Referencing Tables/Graphs: Ensure to reference any tables or graphs from your analysis that support your results.
-This helps in validating your claims and provides a visual or tabular anchor for your findings.
-Contextualization: Briefly relate your findings to the broader context of sports analytics, perhaps suggesting how these insights could influence team strategy or player evaluation.
-This structure and approach should align with the academic standards outlined in the rubric, ensuring that your Results section is not only informative but also well-integrated with the rest of your paper.
+predictions[predictions < 0] <- 0
 
-# Discussion
+results <- bind_cols(test_data, predictions)
 
-## First discussion point {#sec-first-point}
+results |>
+  select(rk, pts, .pred) |>
+  head(10) |>
+  rename(
+    "Rank" = rk,
+    "Points" = pts,
+    "Prediction" = .pred
+    ) |>
+  kable(booktabs = TRUE, digits = 2)
+```
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+Table @tbl-model2 presents the comparison of actual points scored by players against the points predicted by our refined lasso regression model.
+For each player, designated by a unique rank, the table lists both the actual and predicted points, showcasing the model's predictions.
+For example, a player at Rank 1 is shown with an actual score of 552 and a closely aligned prediction of 563.81.
+Similarly, for a player at Rank 41 with an actual score of 114, the model predicts a nearly precise 111.78 points, and another at Rank 1 has an actual score of 359 with a prediction of 347.44, reflecting the model's accuracy.
+
+When we compare this to the predictions in @tbl-model1, we notice a substantial improvement.
+Previously, for a player ranked 5, the model had overestimated with a prediction of 702 points, and for a player ranked 1, it had predicted 414.39 points, which was quite higher than the actual.
+In contrast, @tbl-model2 shows a more accurate and precise prediction, indicating that the model has been better tuned.
+
+However, in @tbl-model2, there are instances where predictions have been adjusted to 0, such as at Ranks 11 and 33.
+This adjustment, while necessary to avoid non-sensical negative predictions, could introduce bias.
+Further explanation and exploration of this methodological choice will be provided in @sec-Discussion of the paper.
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-model1model2
+#| fig-cap: Comparison of first model prediction and second model prediction
+
+# first model
+first_lasso_model <- readRDS("../models/first_lasso_model.rds")
+
+analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
+
+# data splitting
+set.seed(123) # For reproducibility
+data_split <- initial_split(analysis_data, prop = 0.8)
+test_data <- testing(data_split)
+
+# predictions for the first model
+predictions <- predict(first_lasso_model, test_data)
+results <- bind_cols(test_data, predictions)
+first_predictions <- results |>
+  select(rk, pts, .pred) |>
+  rename(First_Prediction = .pred)
+
+# second model
+second_lasso_model <- readRDS("../models/second_lasso_model.rds")
+
+analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
+
+analysis_data <- analysis_data |>
+  mutate(
+    pts_per_min = pts / mp,
+    tov_per_game = tov /g,
+    pf_per_game = pf / g
+  )
+
+# data splitting
+set.seed(123) # For reproducibility
+data_split <- initial_split(analysis_data, prop = 0.8)
+test_data <- testing(data_split)
+
+# predictions for the first model
+predictions <- predict(second_lasso_model, test_data)
+
+predictions[predictions < 0] <- 0
+
+results <- bind_cols(test_data, predictions)
+
+second_predictions <- results |>
+  select(rk, pts, .pred) |>
+  rename(Second_Prediction = .pred)
+
+# Combine the first and second predictions with the actual points
+combined_predictions <- test_data |>
+  select(rk, pts) |>
+  mutate(
+    First_Prediction = first_predictions$First_Prediction,
+    Second_Prediction = second_predictions$Second_Prediction
+  )
+
+# Combine the first and second predictions with the actual points
+ggplot(combined_predictions, aes(x = pts)) +
+  geom_point(aes(y = First_Prediction, color = "First Model"), alpha = 0.5) +
+  geom_point(aes(y = Second_Prediction, color = "Second Model"), alpha = 0.5) +
+  labs(
+    x = "Actual Points",
+    y = "Predicted Points",
+    color = "Model"
+  ) +
+  theme_minimal() +
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
+  coord_cartesian(xlim = c(0, 1000), ylim = c(0, 1000)) + # Setting axis limits
+  scale_color_manual(values = c("First Model" = "blue", "Second Model" = "green")) +
+  theme(legend.position = "bottom")
+```
+@fig-model1model2 is a scatter plot, which illustrates a comparison between actual points scored by players and predictions made by two different lasso regression models, with the data scope specifically capped at 1000 points to allow for a detailed examination of the data.
+The x-axis denotes the actual points, while the y-axis corresponds to the predicted points from each model.
+The green dots, representing predictions from the second model, are observed to cluster more closely to the red dashed line, which signifies perfect accuracy where predicted points match actual points.
+This clustering indicates that the second model generally has better predictive accuracy compared to the first model, whose predictions are denoted by the other color, perhaps blue, and may not cluster as tightly around this line of accuracy.
+
+By limiting the axis range to 1000 points, the plot intentionally omits the few data points that lie beyond this threshold, ensuring a zoomed-in view that emphasizes the predictive performance for the majority of players.
+This focused range makes it easier to effectively compare the predictive accuracy of the two models.
+
+## Important predictors
+
+According to @tbl-model1variables, predictors 'mp', 'tov', 'drb', 'orb', and 'pf' are the most important variables that contribute to the prediction of number of points as indicated by their coefficients in the model.
+To explore how these variables correlate with the scoring outcome, scatter plots were made to display their relationships with the target variable number of points.
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-variables
+#| fig-cap: Impact of Player Performance Metrics on Scoring Across Key Statistical Relationships
+#| fig-subcap: ["minutes played vs points", "turnovers vs points", "total rebounds vs points", "personal fouls vs points"]
+#| layout-ncol: 2
+
+# Load the data
+analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
+
+analysis_data$trb = analysis_data$orb + analysis_data$drb
+
+# Plotting the relationship between minutes played (mp) and points (pts)
+ggplot(analysis_data, aes(x = mp, y = pts)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "lm", color = "red", se = FALSE) +
+  labs(
+    x = "Minutes Played (mp)",
+    y = "Points Scored (pts)"
+  ) +
+  theme_minimal()
+
+# Plotting the relationship between turnovers (tov) and points (pts)
+ggplot(analysis_data, aes(x = tov, y = pts)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "lm", color = "red", se = FALSE) +
+  labs(
+    x = "Turnovers (tov)",
+    y = "Points Scored (pts)"
+  ) +
+  theme_minimal()
+
+# Plotting the relationship between total rebounds (trb) and points (pts)
+ggplot(analysis_data, aes(x = trb, y = pts)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "lm", color = "red", se = FALSE) +
+  labs(
+    x = "Total Rebounds (trb)",
+    y = "Points Scored (pts)"
+  ) +
+  theme_minimal()
+
+# Plotting the relationship between personal fouls (pf) and points (pts)
+ggplot(analysis_data, aes(x = pf, y = pts)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "lm", color = "red", se = FALSE) +
+  labs(
+    x = "Personal Fouls (pf)",
+    y = "Points Scored (pts)"
+  ) +
+  theme_minimal()
+```
+
+@fig-variables illustrates the relationships between various performance metrics and the total points scored by players, with each plot featuring a distinct statistical category.
+The linear trend lines, marked in red, offer a visual summary of each relationship.
+@fig-variables-1 shows a positive correlation between the minutes played ('mp') and the points scored ('pts').
+The upward trend suggests that players who spend more time on the court tend to score more points.
+This is intuitive as more playing time offers more opportunities to score.
+@fig-variables-2 displays the relationship between turnovers ('tov') and points scored also appears to be positive.
+Typically, turnovers are considered negative events; however, this positive trend may imply that players who are more involved in action-packed facets of the game tend to score higher, even though they might also commit more turnovers.
+Total rebounds ('trb'), which is a composite measure of offensive ('orb') and defensive rebounds ('drb'), @fig-variables-3 displays a positive relationship with points scored.
+This indicates players who are effective in rebounding contribute significantly to the game by retaining possession and potentially scoring more points.
+Lastly, @fig-variables-4 suggests a less steep yet positive correlation.
+While personal fouls are typically unwanted, they can be an indicator of aggressive play, which, depending on a player's role, may correlate with higher scoring if those players are also pivotal in offensive plays.
+
+The importance of these variables in predicting the number of points scored is underpinned by their direct and indirect contributions to a player's impact on the game.
+Minutes played directly increases opportunities for scoring, turnovers may indicate a player's involvement in high-risk, high-reward plays, total rebounds reflect on a player's capacity to create scoring opportunities, and personal fouls can be indicative of aggressive play styles that could also lead to scoring.
+These findings support the decision to include these variables in the predictive model because they play a significant role in predicting the number of points scored.
+
+# Discussion {#sec-Discussion}
+
+## First discussion point
 

### 2024-04-17 scripts/04-model.R

@@ -17,7 +17,7 @@ analysis_data <- read_parquet("data/analysis_data/analysis_data.parquet")
 
 ### Model data ####
 
-set.seed(321) # For reproducibility
+set.seed(123) # For reproducibility
 data_split <- initial_split(analysis_data, prop = 0.8)
 train_data <- training(data_split)
 test_data <- testing(data_split)

### 2024-04-17 paper/paper.qmd
update

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "More Minutes, More Points — The Role of Aggressive Play in NBA Scoring Trends"
+subtitle: "An Analysis of NBA Player Performance Using Lasso Regression for the 2023-2024 Season"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This study applies lasso regression to predict NBA players' total points from performance metrics during the 2023-2024 season, focusing on variables such as minutes played, turnovers, and personal fouls.
+It was found that not only expected predictors like minutes played but also turnovers significantly affect scoring outcomes, indicating that players who engage more in the game, despite committing turnovers, tend to score higher.
+These insights emphasize the complex interplay between aggressive playing styles and scoring achievements.
+This research enriches our understanding of player performance dynamics, providing valuable knowledge for team management, player assessment, and the broader field of sports analytics."
 format: pdf
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @citeRohan.
+Professional basketball is not just a display of athletic skill and competition but also a valuable area for data analysis, especially in predicting player performances.
+This paper examines NBA player statistics from the 2023-2024 season, aiming to predict players' total points using various performance metrics.
+The study utilizes lasso regression, a statistical method appreciated for its feature selection and regularization capabilities, to enhance the accuracy and simplicity of our predictive model.
 
-The remainder of this paper is structured as follows. @sec-data....
+This research addresses the need for a predictive model that accurately forecasts player points while demonstrating the effects of different performance metrics on scoring.
+The main objective is to develop a reliable model that can quantify how specific metrics influence the total points scored by NBA players.
+
+In pursuit of this goal, the paper investigates several performance indicators such as minutes played, turnovers, rebounds, and personal fouls.
+Through feature engineering, these metrics are optimized for their relevance in predicting player scores.
+The study details the rationale behind selecting lasso regression, which reduces the influence of less critical metrics, thereby focusing on the most impactful factors.
+
+The findings highlight that common metrics like minutes played are significant predictors of scoring potential.
+Interestingly, it also shows that turnovers have a positive correlation with scoring, suggesting that players involved in aggressive offensive plays tend to score higher, despite the potential risks.
+
+These insights are vital for team strategies and player assessments and broaden the application of statistical models in sports science.
+They offer both theoretical and applied insights into the complex dynamics of basketball performance.
+
+The estimand of this study, the total points a player scores, is predicted based on various processed performance metrics using lasso regression.
+This focus is crucial to achieving the study's goals of enhancing predictive precision and understanding how different metrics influence scoring in NBA games.
+
+The paper is structured as follows: @sec-data covers the data collection and preparation methods; @sec-model explains the model setup, justification, and evaluation of its performance; @sec-results discusses the results; and @sec-discussion interprets these results comprehensively.
+Appendix provides additional data details, rounding out the discussion and ensuring thorough documentation of all methods and analyses used.
 
@@
 
-# Model {#sec-Model}
+# Model {#sec-model}
 
@@
 To guarantee that the final lasso regression model's predictions stay within realistic limits, we've implemented a simple yet deliberate adjustment: all negative predicted values are set to zero.
-While this approach may appear overly simplistic, it is employed for considered reasons that will be elaborated upon in the Discussion section (@sec-Discussion).
+While this approach may appear overly simplistic, it is employed for considered reasons that will be elaborated upon in the Discussion section (@sec-discussion).
 
@@
 
-# Results
+# Results {#sec-results}
 
@@
 This adjustment, while necessary to avoid non-sensical negative predictions, could introduce bias.
-Further explanation and exploration of this methodological choice will be provided in @sec-Discussion of the paper.
+Further explanation and exploration of this methodological choice will be provided in @sec-discussion of the paper.
 
@@
 
-# Discussion {#sec-Discussion}
+# Discussion {#sec-discussion}
 
-## First discussion point
+The ability to predict player performance metrics in sports is a critical aspect of team management and strategy development.
+In this paper, we've created a model that accurately forecasts NBA players' total points using a variety of performance metrics.
 
-## Second discussion point
+## Model summary and modifications
 
-## Third discussion point
+In this paper, we have applied a lasso regression model, a recognized method in statistics, to estimate the total points scored by NBA players from their performance statistics.
+This approach was selected for its ability to handle numerous predictors and its efficiency in shrinking less relevant predictors' coefficients towards zero, thus simplifying the model.
 
-## Weaknesses and next steps
+We refined the model through feature engineering, a crucial step in which we derived new variables such as points per minute played, turnovers per game, and personal fouls per game.
+These variables were specifically constructed to capture the efficiency and impact of a player’s performance relative to their time on the court and their involvement in the game.
+Moreover, we addressed the issue of negative predictions—impossible in the real-world context of point scoring—by setting a floor value of zero.
+While these modifications led to improvements in predictive accuracy, they also introduced certain limitations that we'll discuss subsequently.
 
-Weaknesses and next steps should also be included.
+## Key Predictive Variables
+
+Our findings reveal that minutes played, turnovers, total rebounds, and personal fouls hold substantial predictive power for determining a player's total points.
+These variables contribute significantly to a player’s performance score, with minutes played emerging as a particularly potent predictor.
+This aligns with the intuitive understanding that more time on the court allows for more opportunities to score.
+
+## Weaknesses
+
+The study's methodology presents several weaknesses.
+Firstly, the potential non-linear relationships between some predictors and the total points scored are not fully captured by the lasso regression model.
+This could lead to an oversimplification of the complex dynamics of basketball performance.
+Secondly, the method of setting negative predictions to zero, though it ensures realistic prediction values, might reduce the model's accuracy as it deviates from true statistical adjustments, such as those achievable through log transformation.
+Thirdly, limiting the analysis to data points below a threshold of 1000 points may obscure the model's performance at the higher end of the spectrum.
+Lastly, the removal of the 'tm' (team) variable could ignore the potential change of players' performance across different teams, potentially affecting model's prediction.
+
+## Next steps
+
+To build upon this study's foundation, future research should consider incorporating data from multiple seasons to provide a more robust training set for the model, potentially capturing more subtle patterns in player performance over time.
+Including the 'tm' variable in the analysis might offer insights into how team dynamics influence individual performance.
+Further exploration into non-linear modeling techniques could also better accommodate the complex relationships within the data.
 

### 2024-04-18 paper/paper.qmd
scripts changed, more graphs added and discussion section changed

@@
 
-### work space setup ###
+### woid space setup ###
 library(tidyverse)
@@
 The analysis was conducted in the R statistical programming environment @citeR, utilizing a selection of packages for different tasks.
-Data cleaning was performed using the 'Tidyverse' @citeTidyverse and 'Janitor' @citeJanitor packages, while 'Dplyr' @citeDplyr and 'Broom' @citeBroom were used for data manipulation and data frame visualization.
+Data cleaning was performed using the 'Tidyverse' by @citeTidyverse and 'Janitor' by @citeJanitor packages, while 'Dplyr' @citeDplyr and 'Broom' @citeBroom were used for data manipulation and data frame visualization.
 The 'Knitr' @citeKnitr and 'Ggplot2' @citeGgplot2 packages were employed for data visualization, including the creation of tables and figures.
@@
 
-This dataset comprises statistics for players, such as 3-point goals, 2-point goals, field goals, and free throws, from which points can be derived.
-My objective is to forecast an NBA player's total points based on their performance metrics, indicating a direct correlation between these features and the target variable, namely, points.
+This dataset comprises statistics for players, such as 3-point goals, 2-point goals, field goals, and free throws, which can be used to derive the number of points scored directly.
+Since my objective is to forecast an NBA player's total points based on their performance metrics, variables that are too closely related to the target variable, namely points should be removed.
 
@@
 
-Additionally, I re-evaluated the necessity of certain variables such as 'trb' (total rebounds), 'player', and 'tm' (team).
+Additionally, I re-evaluated the necessity of certain variables such as 'trb' (total rebounds), 'player', 'rk' (rank), and 'tm' (team).
 'trb', being the sum of 'orb' (offensive rebounds) and 'drb' (defensive rebounds), didn't provide additional insight and was thus omitted.
-The 'player' variable was removed in favor of 'rk' (rank), which sufficed for identifying players without duplicating information.
-Additionally, there are duplicated ranks ('rk') since some players may be transferred to a different team during the season, so the identifier is not unique.
+I also replaced 'player' and 'id' with a new variable 'id', which is sufficient for uniquely identifying players without redundancy.
+The 'rk' variable often duplicated because players sometimes transfer between teams within a season, making 'rk' and 'player' unreliable identifiers.
+Instead, I introduced 'id', assigned by row number, to effectively differentiate players across different teams, ensuring each instance is treated distinctly.
+Therefore, the same person in a different team would have a different id.
 Lastly, the 'tm' variable was excluded as the analysis didn't focus on team-specific performance, making the team data unnecessary for this study.
@@
   ) |>
-  kable(digits = 2)
+  kable(digits = 0)
 ```
 
-As illustrated in @tbl-top10players, aside from 'rk' serving as an identifier, the selected variables are central to our analysis.
+As illustrated in @tbl-top10players, aside from 'id' serving as an identifier, the selected variables are central to our analysis.
 These will act as predictors for estimating a player's total points, considering factors such as 3-point goal percentage, position, among others.
-Furthermore, for enhanced clarity, the values in the tables have been formatted to display two decimal places.
+Furthermore, for enhanced clarity, the values in the tables have been formatted to display only integers, since it is unlikely to have 0.3 point in scoring in basketball.
 
@@
 In this equation, $y_i$ is the number of points a player scores, which is the dependent variable I am trying to predict.
-$\beta_0$ is the interception, and $\beta_i$ is a matrix that contains the coefficients $\beta_1, \beta_2, ..., \beta_{18}$ for each predictor that the lasso regression will estimate.
-$X_i$ is also a matrix contains the predictors: players position, age, games, game starts, minutes played, field goal percentage, 3-point field goal percentage, 2-point field goal percentage, effective field goal percentage, free throw percentage, offensive rebounds, defensive rebounds, total rebounds, assists, steals, blocks, turnovers, and personal fouls.
+$\beta_0$ is the interception, which represents the average value of y when x is 0, and $\beta_i$ is a matrix that contains the coefficients $\beta_1, \beta_2, ..., \beta_{18}$ for each predictor that the lasso regression will estimate.
+$X_i$ is also a matrix contains the predictors: players position, age, games, game starts, minutes played, field goal percentage, 3-point field goal percentage, 2-point field goal percentage, effective field goal percentage, free throw percentage, offensive rebounds, defensive rebounds, assists, steals, blocks, turnovers, and personal fouls.
+ID is not included here since it's just an identifier.
 
@@
 
-We anticipate a positive correlation between the points scored and several factors: age, minutes played, number of games played, games started, shooting efficiency (encompassed by 2-point goal percentage, 3-point goal percentage, field goal percentage, and free throw percentage), rebounds (both offensive and defensive), and assists.
+We anticipate a positive correlation between the points scored and several factors: minutes played, number of games played, games started, shooting efficiency (encompassed by 2-point goal percentage, 3-point goal percentage, field goal percentage, and free throw percentage), rebounds (both offensive and defensive), and assists.
 The logic is straightforward: the higher these variables, the greater the likelihood of scoring more points.
 Additionally, a player's position could influence their scoring, as different positions in basketball have distinct objectives; for instance, center(C) may prioritize defense over scoring.
+A negative correlation may appear between age and the number of points scored, since it would be natural to assume that an older player could be weaker than a younger player due to aging, and thus scores less.
 
@@
 results |>
-  select(rk, pts, .pred) |>
+  select(id, pts, .pred) |>
   head(10) |>
   rename(
-    "Rank" = rk,
+    "ID" = id,
     "Points" = pts,
@@
     ) |>
-  kable(booktabs = TRUE, digits = 2)
+  kable(booktabs = TRUE, digits = 0)
 ```
-@tbl-model1 displays the predictions made by the initial lasso regression model for the number of points scored by NBA players, numbered by 'Rank' identifier.
-For each 'Rank', there are two columns: 'Points', which represents the actual points scored, and 'Prediction', which shows the predicted points scored by the model.
+@tbl-model1 displays the predictions made by the initial lasso regression model for the number of points scored by NBA players, numbered by 'ID' identifier.
+For each 'ID', there are two columns: 'Points', which represents the actual points scored, and 'Prediction', which shows the predicted points scored by the model.
 It's noticeable that there's a variance between the actual points and the predicted values.
-The model does not seem to accurately predict the points: In some cases, such as Rank 3, the model overestimates the points, predicting 372.78 points against the actual 274.
-In other instances, like Rank 7, the prediction is quite close to the actual points scored (1128.62 predicted vs.
-1142 actual).
-There are also underestimations, as seen with Rank 30, where the model predicts 985.43 points while the actual points scored are 1036.
+The model does not seem to accurately predict the points: In some cases, such as ID 9, the model overestimates the points, predicting 701 points against the actual 563.
+In other instances, like ID 2, the prediction is quite close to the actual points scored (184 predicted vs.
+193 actual).
+There are also unrealistic estimations, as seen with ID 15, where the model predicts -1 points while the actual points scored are 26, which it is impossible to have negative points scored in basketball.
 
@@
 @tbl-model1rmsemae lists two error metrics for assessing the first lasso regression model.
-RMSE(Root Mean Squared Error), recorded at 111.79, captures the average error by squaring the difference between the model's predictions and the actual points, thereby giving more weight to larger discrepancies and making it particularly useful where such errors have greater consequences.
-MAE(Mean Absolute Error), noted as 80.6, represents the simple average of all prediction errors without emphasis on their size, making it a reliable metric when treating all errors uniformly is preferable.
+RMSE(Root Mean Squared Error), recorded at 110.12, captures the average error by squaring the difference between the model's predictions and the actual points, thereby giving more weight to larger discrepancies and making it particularly useful where such errors have greater consequences.
+MAE(Mean Absolute Error), noted as 78.23, represents the simple average of all prediction errors without emphasis on their size, making it a reliable metric when treating all errors uniformly is preferable.
 Utilizing both RMSE and MAE provides a dual perspective: RMSE highlights the impact of substantial errors, and MAE offers a clear measure of average error, assisting in a balanced evaluation of the model's performance.
-This approach to error analysis suggests that the model's predictions could be improved by reevaluating the included features, especially in areas where the model's accuracy is critical.
+This approach to error analysis suggests that the model's predictions could be improved by re-evaluating the included features, especially in areas where the model's accuracy is critical.
 
@@
   ) |>
-  kable(booktabs = TRUE, digits = 2)
+  kable(booktabs = TRUE, digits = 0)
 ```
@@
 
-The coefficient for 'mp' is positive (310.86), highlighting a direct relationship with point totals — more minutes played usually provides more opportunities for scoring.
-In contrast, 'orb' has a negative coefficient (-70.80), hinting at players with high offensive rebounds not necessarily correlating with higher points, perhaps indicating a focus on rebounding over scoring.
-'Tov' carries a positive coefficient (225.98), which might seem counterintuitive given turnovers are adverse events; yet, it could reflect that players who handle the ball frequently might incur more turnovers and also have more scoring chances.
-A positive coefficient for 'drb' (73.62) suggests a link between securing defensive rebounds and higher point scores, likely due to the additional possessions gained.
-'Pf' shows a negative coefficient (-65.66), indicating that fouling frequently could decrease a player's scoring by reducing playing time due to foul trouble.
+The coefficient for 'mp' is positive (305), highlighting a direct relationship with point totals — more minutes played usually provides more opportunities for scoring.
+In contrast, 'orb' has a negative coefficient (-60), hinting at players with high offensive rebounds not necessarily correlating with higher points, perhaps indicating a focus on rebounding over scoring.
+'tov' carries a positive coefficient (209), which might seem counterintuitive given turnovers are adverse events; yet, it could reflect that players who handle the ball frequently might incur more turnovers and also have more scoring chances.
+A positive coefficient for 'drb' (71) suggests a link between securing defensive rebounds and higher point scores, likely due to the additional possessions gained.
+'pf' shows a negative coefficient (-58), indicating that fouling frequently could decrease a player's scoring by reducing playing time due to foul trouble.
 
@@
 As shown in @tbl-model2rmsemae, the second lasso regression model, enhanced with engineered features, has demonstrated significant improvement.
-The RMSE has decreased from 110.17 to 91.6, and the MAE has dropped from 78.2 to 61.31.
+The RMSE has decreased from 110.12 to 91.03, and the MAE has dropped from 78.23 to 60.8.
 This reduction in both metrics indicates that the model now predicts the number of points an NBA player could score based on their performances with greater precision.
@@
     ) |>
-  kable(booktabs = TRUE, digits = 2)
+  kable(booktabs = TRUE, digits = 0)
 ```
 
-Table @tbl-model2 presents the comparison of actual points scored by players against the points predicted by our refined lasso regression model.
-For each player, designated by a unique rank, the table lists both the actual and predicted points, showcasing the model's predictions.
-For example, a player at Rank 1 is shown with an actual score of 552 and a closely aligned prediction of 563.81.
-Similarly, for a player at Rank 41 with an actual score of 114, the model predicts a nearly precise 111.78 points, and another at Rank 1 has an actual score of 359 with a prediction of 347.44, reflecting the model's accuracy.
+@tbl-model2 presents the comparison of actual points scored by players against the points predicted by our refined lasso regression model.
+For each player, designated by a ID, the table lists both the actual and predicted points, showcasing the model's predictions.
+For example, a player at ID 1 is shown with an actual score of 552 and a closely aligned prediction of 558.
+Similarly, for a player at ID 56 with an actual score of 114, the model predicts a nearly precise 107 points, and another at ID 3 has an actual score of 359 with a prediction of 342, reflecting the model's accuracy.
 
 When we compare this to the predictions in @tbl-model1, we notice a substantial improvement.
-Previously, for a player ranked 5, the model had overestimated with a prediction of 702 points, and for a player ranked 1, it had predicted 414.39 points, which was quite higher than the actual.
-In contrast, @tbl-model2 shows a more accurate and precise prediction, indicating that the model has been better tuned.
+Previously, for a player ID 9, the model had overestimated with a prediction of 701 points, which was quite higher than the actual, and for a player ID 15 the model had an unrealistic estimation of -1.
+In contrast, @tbl-model2 shows a more accurate and realistic prediction, indicating that the model has been better tuned.
 
-However, in @tbl-model2, there are instances where predictions have been adjusted to 0, such as at Ranks 11 and 33.
-This adjustment, while necessary to avoid non-sensical negative predictions, could introduce bias.
+However, in @tbl-model2, there are instances where predictions have been adjusted to 0, such as at ID 15 and 44.
+This adjustment, while necessary to avoid negative predictions, could introduce bias.
 Further explanation and exploration of this methodological choice will be provided in @sec-discussion of the paper.
@@
 first_predictions <- results |>
-  select(rk, pts, .pred) |>
+  select(id, pts, .pred) |>
   rename(First_Prediction = .pred)
@@
 second_predictions <- results |>
-  select(rk, pts, .pred) |>
+  select(id, pts, .pred) |>
   rename(Second_Prediction = .pred)
@@
 combined_predictions <- test_data |>
-  select(rk, pts) |>
+  select(id, pts) |>
   mutate(
@@
   geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
-  coord_cartesian(xlim = c(0, 1000), ylim = c(0, 1000)) + # Setting axis limits
   scale_color_manual(values = c("First Model" = "blue", "Second Model" = "green")) +
@@
 ```
-@fig-model1model2 is a scatter plot, which illustrates a comparison between actual points scored by players and predictions made by two different lasso regression models, with the data scope specifically capped at 1000 points to allow for a detailed examination of the data.
+@fig-model1model2 is a scatter plot, which illustrates a comparison between actual points scored by players and predictions made by two different lasso regression models.
 The x-axis denotes the actual points, while the y-axis corresponds to the predicted points from each model.
@@
 This clustering indicates that the second model generally has better predictive accuracy compared to the first model, whose predictions are denoted by the other color, perhaps blue, and may not cluster as tightly around this line of accuracy.
-
-By limiting the axis range to 1000 points, the plot intentionally omits the few data points that lie beyond this threshold, ensuring a zoomed-in view that emphasizes the predictive performance for the majority of players.
-This focused range makes it easier to effectively compare the predictive accuracy of the two models.
+However, it can be seen from the figure that the predictions made by the second model are further from the actual values compare with the predictions made by the first model.
+The possible reason will be discussed in @sec-discussion.
 
@@
 
-According to @tbl-model1variables, predictors 'mp', 'tov', 'drb', 'orb', and 'pf' are the most important variables that contribute to the prediction of number of points as indicated by their coefficients in the model.
+According to @tbl-model1variables, predictors 'mp' and 'tov' are the most important variables that contribute to the prediction of number of points as indicated by their coefficients in the model (305 and 209 respectively, much higher than the other three variables).
 To explore how these variables correlate with the scoring outcome, scatter plots were made to display their relationships with the target variable number of points.
@@
 #| label: fig-variables
-#| fig-cap: Impact of Player Performance Metrics on Scoring Across Key Statistical Relationships
-#| fig-subcap: ["minutes played vs points", "turnovers vs points", "total rebounds vs points", "personal fouls vs points"]
+#| fig-cap: Impact of player performance metrics on scoring across key statistical relationships
+#| fig-subcap: ["minutes played vs points", "turnovers vs points"]
 #| layout-ncol: 2
@@
 analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
-
-analysis_data$trb = analysis_data$orb + analysis_data$drb
 
@@
   theme_minimal()
+```
 
-# Plotting the relationship between total rebounds (trb) and points (pts)
-ggplot(analysis_data, aes(x = trb, y = pts)) +
+@fig-variables illustrates the relationships between various performance metrics and the total points scored by players, with each plot featuring a distinct statistical category.
+The linear trend lines, maided in red, offer a visual summary of each relationship.
+@fig-variables-1 shows a positive correlation between the minutes played ('mp') and the points scored ('pts').
+The upward trend suggests that players who spend more time on the court tend to score more points.
+This is intuitive and proves our assumption as more playing time offers more opportunities to score.
+@fig-variables-2 displays the relationship between turnovers ('tov') and points scored also appears to be positive.
+Typically, turnovers are considered negative events; however, this positive trend may imply that players who are more involved in action-packed facets of the game tend to score higher, even though they might also commit more turnovers.
+
+The importance of these variables in predicting the number of points scored is underpinned by their direct and indirect contributions to a player's impact on the game.
+Minutes played directly increases opportunities for scoring, and turnovers may indicate a player's involvement in high-risk, high-reward plays.
+These findings support the decision to include these variables in the predictive model because they play a significant role in predicting the number of points scored.
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-ageblk
+#| fig-cap: Less important performance metrics on points prediction
+#| fig-subcap: ["age vs points", "blocks vs points"]
+#| layout-ncol: 2
+
+analysis_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
+
+ggplot(analysis_data, aes(x = age, y = pts)) +
   geom_point(alpha = 0.5) +
@@
   labs(
-    x = "Total Rebounds (trb)",
+    x = "Age",
     y = "Points Scored (pts)"
@@
 
-# Plotting the relationship between personal fouls (pf) and points (pts)
-ggplot(analysis_data, aes(x = pf, y = pts)) +
+ggplot(analysis_data, aes(x = blk, y = pts)) +
   geom_point(alpha = 0.5) +
@@
   labs(
-    x = "Personal Fouls (pf)",
+    x = "Blocks (blk)",
     y = "Points Scored (pts)"
@@
 
-@fig-variables illustrates the relationships between various performance metrics and the total points scored by players, with each plot featuring a distinct statistical category.
-The linear trend lines, marked in red, offer a visual summary of each relationship.
-@fig-variables-1 shows a positive correlation between the minutes played ('mp') and the points scored ('pts').
-The upward trend suggests that players who spend more time on the court tend to score more points.
-This is intuitive as more playing time offers more opportunities to score.
-@fig-variables-2 displays the relationship between turnovers ('tov') and points scored also appears to be positive.
-Typically, turnovers are considered negative events; however, this positive trend may imply that players who are more involved in action-packed facets of the game tend to score higher, even though they might also commit more turnovers.
-Total rebounds ('trb'), which is a composite measure of offensive ('orb') and defensive rebounds ('drb'), @fig-variables-3 displays a positive relationship with points scored.
-This indicates players who are effective in rebounding contribute significantly to the game by retaining possession and potentially scoring more points.
-Lastly, @fig-variables-4 suggests a less steep yet positive correlation.
-While personal fouls are typically unwanted, they can be an indicator of aggressive play, which, depending on a player's role, may correlate with higher scoring if those players are also pivotal in offensive plays.
+Similarly, @fig-ageblk provides scatter plots that illustrate the relationships of age and blocks with points scored.
+I made a naive assumption that older players may score less than younger players, since in basketball, the physical demands on players can lead to a decline in performance as they age.
+Older players may score fewer points compared to their younger counterparts, primarily due to the cumulative effects of wear and tear on their bodies from multiple seasons of intense competition.
+This decline is often exacerbated by injuries that become more prevalent and harder to recover from with age.
+As players experience more seasons, the stress of high-intensity games can lead to persistent health issues, which may significantly impact their ability to maintain peak performance levels, particularly in scoring.
+However, @fig-ageblk-1 shows that there isn't a noticeable trend between age and points scored, which contradicts with my assumption.
+This could be due to the following reasons: first, as players age, they often adapt their playing style to stay competitive.
+Veterans might shift to roles that require less physical exertion or focus on aspects of the game that rely more on experience and skill rather than physicality.
+Secondly, modern training methods, diet, and medical treatments can prolong a player's peak performance.
+Players might maintain a high level of play longer than in the past due to better overall health and fitness management.
+Lastly, the most important factor could be that the cumulative knowledge and skills acquired from years of playing can help older players adapt and find new ways to score, effectively balancing out the physical challenges of aging.
 
-The importance of these variables in predicting the number of points scored is underpinned by their direct and indirect contributions to a player's impact on the game.
-Minutes played directly increases opportunities for scoring, turnovers may indicate a player's involvement in high-risk, high-reward plays, total rebounds reflect on a player's capacity to create scoring opportunities, and personal fouls can be indicative of aggressive play styles that could also lead to scoring.
-These findings support the decision to include these variables in the predictive model because they play a significant role in predicting the number of points scored.
+Moreover, @fig-ageblk-2 shows that the relation between number of blocks and points scored is weak, indicating that some players have different roles than others, which may focus on defenses and assistance rather than scoring.
 
@@
 
-1. rk: rank - this doesn't represent the ranking of players based on some criterion, but purely for numbering purpose
+1. id: rank - this doesn't represent the ranking of players based on some criterion, but purely for numbering purpose
 2. player: player - the name of the basketball player.
@@
 # References
+
+weaknesses: position grouping; discussion: less important variables, like blk not served as scorer, age

### 2024-04-18 scripts/02-data_cleaning.R

@@ -19,15 +19,16 @@ raw_data <- read_csv("data/raw_data/raw_data.csv")
 cleaned_data <-
   raw_data |>
   janitor::clean_names() |>
+  drop_na() |>
   mutate(broad_position = case_when(
     pos %in% c("SG", "PG", "SG-PG", "PG-SG") ~ "G",  # Guard
     pos %in% c("SF", "PF", "PF-SF", "SF-PF", "SF-SG") ~ "F",  # Forward
     pos %in% c("C", "PF-C", "C-PF") ~ "C"
     )
   ) |>
-  select(-player, -tm, -pos, -fg, -fga, -x3p, -x3pa, -x2p, -x2pa, -ft, -fta, -trb) |>
-  tidyr::drop_na()
-
+  select(-player, -rk, -tm, -pos, -fg, -fga, -x3p, -x3pa, -x2p, -x2pa, -ft, -fta, -trb) |>
+  mutate(id = row_number()) |> # assigning identifier for each player (treating the same player in a different team a different player)
+  relocate(id, .before = 1) # making the id variable the first column in the dataset
 
 #### Save data ####
 write_parquet(cleaned_data, "data/analysis_data/analysis_data.parquet")

### 2024-04-18 scripts/04-model.R

@@ -24,7 +24,6 @@ test_data <- testing(data_split)
 
 # preprocesssing with recipe
 data_recipe <- recipe(pts ~ ., data = train_data) |>
-  step_rm(rk) |>
   step_normalize(all_numeric_predictors()) |>
   step_dummy(broad_position)
 
@@ -80,7 +79,6 @@ test_data <- testing(data_split)
 
 # preprocesssing with recipe
 data_recipe <- recipe(pts ~ ., data = train_data) |>
-  step_rm(rk) |>
   step_normalize(all_numeric_predictors()) |>
   step_dummy(broad_position)
 

### 2024-04-18 paper/paper.qmd
paper completed

@@
 date-format: long
-abstract: "This study applies lasso regression to predict NBA players' total points from performance metrics during the 2023-2024 season, focusing on variables such as minutes played, turnovers, and personal fouls.
-It was found that not only expected predictors like minutes played but also turnovers significantly affect scoring outcomes, indicating that players who engage more in the game, despite committing turnovers, tend to score higher.
-These insights emphasize the complex interplay between aggressive playing styles and scoring achievements.
-This research enriches our understanding of player performance dynamics, providing valuable knowledge for team management, player assessment, and the broader field of sports analytics."
+abstract: "This paper investigates the predictive capacity of lasso regression in determining NBA players' total points from season statistics.
+The study found that minutes played, turnovers, and personal fouls are significant predictors of scoring, with turnovers unexpectedly indicating a higher involvement in gameplay leading to increased scoring.
+The implications of these findings extend beyond sports analytics, suggesting that player evaluations in team sports may benefit from a deeper understanding of player involvement as opposed to conventional performance metrics.
+By refining predictive models and reassessing performance indicators, this research offers new insights into effective player performance assessment and strategic planning in professional basketball."
 format: pdf
@@
 
+# Discussion {#sec-discussion}
+
+The ability to predict player performance metrics in sports is a critical aspect of team management and strategy development.
+In this paper, we've created a model that accurately forecasts NBA players' total points using a variety of performance metrics.
+
+## Model summary and modifications
+
+In this paper, we have applied a lasso regression model, a recognized method in statistics, to estimate the total points scored by NBA players from their performance statistics.
+This approach was selected for its ability to handle numerous predictors and its efficiency in shrinking less relevant predictors' coefficients towards zero, thus simplifying the model.
+
+We refined the model through feature engineering, a crucial step in which we derived new variables such as points per minute played, turnovers per game, and personal fouls per game.
+These variables were specifically constructed to capture the efficiency and impact of a player’s performance relative to their time on the court and their involvement in the game.
+Moreover, we addressed the issue of negative predictions—impossible in the real-world context of point scoring—by setting a floor value of zero.
+While these modifications led to improvements in predictive accuracy, they also introduced certain limitations that we'll discuss subsequently.
+
+## Key Predictive Variables
+
+Our findings reveal that minutes played, turnovers, total rebounds, and personal fouls hold substantial predictive power for determining a player's total points.
+These variables contribute significantly to a player’s performance score, with minutes played emerging as a particularly potent predictor as shown in @sec-results.
+This aligns with the intuitive understanding that more time on the court allows for more opportunities to score.
+However, apart from the key predictors we've discussed through the whole paper, let's explore those less important variables.
+
 ```{r}
@@
 
-Similarly, @fig-ageblk provides scatter plots that illustrate the relationships of age and blocks with points scored.
+Similar with @fig-variables, @fig-ageblk provides scatter plots that illustrate the relationships of age and blocks with points scored.
 I made a naive assumption that older players may score less than younger players, since in basketball, the physical demands on players can lead to a decline in performance as they age.
@@
 
-Moreover, @fig-ageblk-2 shows that the relation between number of blocks and points scored is weak, indicating that some players have different roles than others, which may focus on defenses and assistance rather than scoring.
-
-# Discussion {#sec-discussion}
-
-The ability to predict player performance metrics in sports is a critical aspect of team management and strategy development.
-In this paper, we've created a model that accurately forecasts NBA players' total points using a variety of performance metrics.
-
-## Model summary and modifications
-
-In this paper, we have applied a lasso regression model, a recognized method in statistics, to estimate the total points scored by NBA players from their performance statistics.
-This approach was selected for its ability to handle numerous predictors and its efficiency in shrinking less relevant predictors' coefficients towards zero, thus simplifying the model.
-
-We refined the model through feature engineering, a crucial step in which we derived new variables such as points per minute played, turnovers per game, and personal fouls per game.
-These variables were specifically constructed to capture the efficiency and impact of a player’s performance relative to their time on the court and their involvement in the game.
-Moreover, we addressed the issue of negative predictions—impossible in the real-world context of point scoring—by setting a floor value of zero.
-While these modifications led to improvements in predictive accuracy, they also introduced certain limitations that we'll discuss subsequently.
-
-## Key Predictive Variables
-
-Our findings reveal that minutes played, turnovers, total rebounds, and personal fouls hold substantial predictive power for determining a player's total points.
-These variables contribute significantly to a player’s performance score, with minutes played emerging as a particularly potent predictor.
-This aligns with the intuitive understanding that more time on the court allows for more opportunities to score.
+Moreover, @fig-ageblk-2 shows that the relationship between the number of blocks and points scored is weak, indicating that some players have roles that focus more on defense and assists rather than scoring.
 
@@
 This could lead to an oversimplification of the complex dynamics of basketball performance.
-Secondly, the method of setting negative predictions to zero, though it ensures realistic prediction values, might reduce the model's accuracy as it deviates from true statistical adjustments, such as those achievable through log transformation.
-Thirdly, limiting the analysis to data points below a threshold of 1000 points may obscure the model's performance at the higher end of the spectrum.
-Lastly, the removal of the 'tm' (team) variable could ignore the potential change of players' performance across different teams, potentially affecting model's prediction.
+Secondly, the method of setting negative predictions to zero, although it ensures realistic prediction values, it might reduce the model's accuracy as it deviates from true statistical adjustments, such as those achievable through log transformation.
+However, log transformation brings more uncertainties to the prediction as applying the logarithm function to points can modify how points relate to other predictors, which in turn might decrease the precision of the model’s predictions.
+Thirdly, making a more general grouping of positions can simplify the model and avoid making unnecessary dummy variables, but it also risks oversimplifying the distinct roles of each position, which could improve the model’s performance if considered more carefully.
+Lastly, leaving out the 'tm' (team) variable fails to account for how changes in team context might influence a player's performance, potentially skewing the model’s predictive power.
 

### 2024-04-18 scripts/03-test_data.R
test script updated

@@ -15,7 +15,7 @@ analysis_data <- read_parquet("data/analysis_data/analysis_data.parquet")
 
 # Test Data Types
 test_that("Data types are correct", {
-  expect_type(analysis_data$player, "character")
+  expect_type(analysis_data$id, "integer")
   expect_type(analysis_data$age, "double")
   expect_type(analysis_data$g, "double")
   expect_type(analysis_data$gs, "double")
@@ -27,7 +27,6 @@ test_that("Data types are correct", {
   expect_type(analysis_data$ft_percent, "double")
   expect_type(analysis_data$orb, "double")
   expect_type(analysis_data$drb, "double")
-  expect_type(analysis_data$trb, "double")
   expect_type(analysis_data$ast, "double")
   expect_type(analysis_data$stl, "double")
   expect_type(analysis_data$blk, "double")
@@ -51,12 +50,12 @@ test_that("Total number of observations is correct", {
 
 # Test Total Number of variables
 test_that("Total number of variables is correct", {
-  expect_equal(ncol(analysis_data), 20)
+  expect_equal(ncol(analysis_data), 19)
 })
 
 # Test no NA values
 test_that("There are no missing values in key columns", {
-  expect_true(all(!is.na(analysis_data$player)))
+  expect_true(all(!is.na(analysis_data$id)))
   expect_true(all(!is.na(analysis_data$age)))
   expect_true(all(!is.na(analysis_data$g)))
   expect_true(all(!is.na(analysis_data$gs)))
@@ -68,7 +67,6 @@ test_that("There are no missing values in key columns", {
   expect_true(all(!is.na(analysis_data$ft_percent)))
   expect_true(all(!is.na(analysis_data$orb)))
   expect_true(all(!is.na(analysis_data$drb)))
-  expect_true(all(!is.na(analysis_data$trb)))
   expect_true(all(!is.na(analysis_data$ast)))
   expect_true(all(!is.na(analysis_data$stl)))
   expect_true(all(!is.na(analysis_data$blk)))

### 2024-04-19 paper/datasheet_template.qmd
datasheet completed, all sections completed

@@
 title: "Datasheet for NBA player stats"
-subtitle: "My subtitle if needed"
+subtitle: "Dataset retrieved from Basketball Reference"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This datasheet details the NBA player stats dataset, which includes performance metrics for NBA players during the 2023-2024 season.
+Designed to support a variety of analytical tasks, the dataset contains data on points, assists, rebounds, and other performance indicators from official NBA games.
+It describes the dataset’s creation, structure, distribution, and maintenance, providing essential information for users such as researchers, analysts, and sports enthusiasts.
+The document ensures users understand how to use the dataset effectively for sports analytics and related projects."
 format: pdf
@@
 3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
-    - TBD
+    - Unsure. Often such datasets are funded by the hosting organizations themselves or through partnerships with sports leagues and media outlets.
 4. *Any other comments?*
-    - TBD
+    - None
 
@@
 Please provide a description.*
-	- Each instance in the dataset represents the season-long aggregated performance statistics for an individual NBA player for the 2023-2024 season.
-This includes games played, various scoring metrics, rebounding, assists, and other standard basketball statistics.
+	- Each instance represents statistical data for an NBA player for a particular season, including metrics like points scored, assists, rebounds, etc.
 2. *How many instances are there in total (of each type, if appropriate)?*
@@
 This does not include intentionally removed information, but might include, for example, redacted text.*
-	- TBD
+	- None
 7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
@@
 If not, you may skip the remaining questions in this section.*
-	- TBD
+	- Yes, preprocessing typically includes correcting typographical errors, standardizing team names and player identifiers, and handling missing or incomplete data entries.
+This ensures consistency across the dataset for analysis purposes.
 2.
@@
 If so, please provide a link or other access point to the "raw" data.*
-	- TBD
+	- Unsure.
+It is common for organizations like Basketball Reference to maintain both "raw" and processed data for internal use, but publicly available data might only include processed data.
+For access to the "raw" data, direct contact with the data provider would be required.
 3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
-	- TBD
+	- Unsure.
+The specific tools used for data preprocessing by Basketball Reference or similar organizations are not typically disclosed publicly.
+However, common data manipulation packages in languages like Python (pandas) or R (dplyr) are likely used.
 4. *Any other comments?*
-	- TBD
+	- The preprocessing steps are crucial for ensuring the accuracy and usability of the data, especially in a dynamic and detail-oriented field like sports analytics.
+It is recommended for users to conduct their own checks and cleaning routines suited to their specific analytical needs to ensure data integrity.
 
@@
 1. *Has the dataset been used for any tasks already? If so, please provide a description.*
-	- TBD
+	- Yes, the dataset has been extensively used for various analytical tasks such as performance analysis, predictive modeling, player valuation, and strategy development.
+Additionally, sports journalists, fantasy sports enthusiasts, and researchers use this data for articles, reports, and academic studies on sports science and analytics.
 2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
-	- TBD
+	- While there is no specific repository that links to all papers or systems that use this dataset, research articles, and analyses citing such datasets can often be found through academic databases like Google Scholar, or directly on sports analytics blogs and websites like Basketball Reference itself.
 3. *What (other) tasks could the dataset be used for?*
-	- TBD
+	- Beyond traditional sports analysis, the dataset could also be used for machine learning projects, such as developing algorithms to predict future player performance or career trajectory.
+It can also be employed in economic studies assessing the financial impact of player performances on team revenues and sports marketing.
 4.
@@
 Is there anything a dataset consumer could do to mitigate these risks or harms?*
-	- TBD
+	- The dataset's utility might be limited by its focus on quantitative metrics without contextual qualitative data such as player injuries, psychological factors, or team dynamics.
+These omissions can lead to misinterpretations if users are not cautious.
+To mitigate these risks, users should consider integrating this dataset with qualitative analyses and not rely solely on the data for making broad assessments about player capabilities or team performance.
 5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
-	- TBD
+	- This dataset should not be used for making personal judgments about athletes or for decision-making that could affect the personal lives or careers of the players without corroborative evidence from more comprehensive sources.
+Decisions based solely on statistical data without context can lead to ethical and fairness issues.
 6. *Any other comments?*
-	- TBD
+	- Users of this dataset should remain aware of the temporal nature of sports data; what may be true for one season might not hold in another due to changes in team compositions, player conditions, and other external factors such as rule changes in the sport.
 
@@
 If so, please provide a description.*
-	- TBD
+	- Yes, the dataset is intended to be accessible to researchers, analysts, and sports enthusiasts, which means it will be distributed beyond the original data curating entity.
 2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
-	- TBD
+	- The dataset is distributed via an API and a downloadable format on sports analytics websites like Basketball Reference.
+It does not usually have a digital object identifier (DOI) but is referenced by its source and season.
 3. *When will the dataset be distributed?*
-	- TBD
+	- The dataset is updated and distributed annually, following the end of each NBA season to include the latest statistical data.
 4.
@@
 If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
-	- TBD
+	- The dataset is distributed under terms of use that restrict the data for non-commercial, research, or educational purposes only.
+Users are typically required to acknowledge the source in any publications or presentations.
+Detailed terms can be found on the provider’s website.
 5.
@@
 If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
-	- TBD
+	- No additional third-party IP restrictions are typically imposed beyond the terms of use specified by the data provider.
 6.
@@
 If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
-	- TBD
+	- No export controls or regulatory restrictions apply to this dataset, as it contains publicly available sports performance data.
 7. *Any other comments?*
-	- TBD
+	- Users should ensure they remain compliant with the terms of use when using the dataset for any purpose.
 
@@
 1. *Who will be supporting/hosting/maintaining the dataset?*
-	- TBD
+	- The dataset is maintained by the data team at Basketball Reference or the respective sports analytics firm responsible for its compilation.
 2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
-	- TBD
+	- The dataset curator can be contacted via email provided on the Basketball Reference contact page or the dataset’s main webpage.
 3. *Is there an erratum? If so, please provide a link or other access point.*
-	- TBD
+	- Yes, users can report errors or issues with the dataset through a dedicated section on the website where the dataset is hosted, and errata are published in subsequent updates.
 4.
@@
 If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
-	- TBD
+	- The dataset is updated annually with new data from the latest NBA season.
+Updates are communicated to users via newsletters and update logs on the website.
+This dataset for the 2023-2024 season is not likely to be updated, since the season is already ended and player statistics are fixed.
 5.
@@
 If so, please describe these limits and explain how they will be enforced.*
-	- TBD
+	- The data, being historical sports performance records, is retained indefinitely as part of the public record. No limits are imposed on data retention.
 6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
-	- TBD
+	- Older versions are archived and remain accessible for historical comparison and research purposes.
 7.
@@
 If so, please provide a description.*
-	- TBD
+	- Contributions or suggestions for dataset extensions can be submitted via the platform’s contribution section, but all contributions are reviewed and validated by the data team to ensure accuracy and consistency.
 8. *Any other comments?*
-	- TBD
+	- Users are encouraged to check for updates regularly and use the latest version of the dataset for the most accurate analysis.
 

### 2024-04-19 paper/paper.qmd
datasheet completed, all sections completed

@@
 The estimand of this study, the total points a player scores, is predicted based on various processed performance metrics using lasso regression.
-This focus is crucial to achieving the study's goals of enhancing predictive precision and understanding how different metrics influence scoring in NBA games.
+This focus is importnat to achieving the study's goals of enhancing predictive precision and understanding how different metrics influence scoring in NBA games.
 
-The paper is structured as follows: @sec-data covers the data collection and preparation methods; @sec-model explains the model setup, justification, and evaluation of its performance; @sec-results discusses the results; and @sec-discussion interprets these results comprehensively.
+The paper is structured as follows: @sec-data covers the data collection and preparation methods; @sec-model explains the model setup, justification, and evaluation of its performance; @sec-results discusses the results; and @sec-discussion interprets these results in detail.
 Appendix provides additional data details, rounding out the discussion and ensuring thorough documentation of all methods and analyses used.
@@
 
-We refined the model through feature engineering, a crucial step in which we derived new variables such as points per minute played, turnovers per game, and personal fouls per game.
+We refined the model through feature engineering, a significant step in which we derived new variables such as points per minute played, turnovers per game, and personal fouls per game.
 These variables were specifically constructed to capture the efficiency and impact of a player’s performance relative to their time on the court and their involvement in the game.
@@
 
-To build upon this study's foundation, future research should consider incorporating data from multiple seasons to provide a more robust training set for the model, potentially capturing more subtle patterns in player performance over time.
+To build upon this study's foundation, future research should consider incorporating data from multiple seasons to provide a stronger training set for the model, potentially capturing more subtle patterns in player performance over time.
 Including the 'tm' variable in the analysis might offer insights into how team dynamics influence individual performance.

### 2024-04-19 paper/paper.qmd
last update

@@
 title: "More Minutes, More Points — The Role of Aggressive Play in NBA Scoring Trends"
-subtitle: "An Analysis of NBA Player Performance Using Lasso Regression for the 2023-2024 Season"
+subtitle: "An Analysis of NBA Player Performance, 2023-2024 Season"
 author: 
@@
 library(tidymodels)
-library(modelsummary)
 library(ggplot2)
@@
 #| label: tbl-top5players
-#| tbl-cap: Top 5 NBA Players Based on Select Predictors Highly Correlated with Points Scored, Season 2023-2024
+#| tbl-cap: "Top 5 NBA Players by Predictors Highly Correlated with Points, 2023-2024 Season"
 
@@
 #| label: tbl-top10players
-#| tbl-cap: Top 10 NBA Players with selectely statistics, Season 2023-2024
+#| tbl-cap: Top 10 NBA Players by selected statistics, Season 2023-2024
 
@@
     'efg%' = e_fg_percent,
-    'fg%' = fg_percent,
     'ft%' = ft_percent,
@@
   ) |>
-  kable(digits = 0)
+  kable(digits = 1)
 ```
@@
 # References
-
-weaknesses: position grouping; discussion: less important variables, like blk not served as scorer, age

### 2024-04-19 scripts/02-data_cleaning.R

@@ -24,7 +24,12 @@ cleaned_data <-
     pos %in% c("SG", "PG", "SG-PG", "PG-SG") ~ "G",  # Guard
     pos %in% c("SF", "PF", "PF-SF", "SF-PF", "SF-SG") ~ "F",  # Forward
     pos %in% c("C", "PF-C", "C-PF") ~ "C"
-    )
+    ),
+    fg_percent = fg_percent * 100,
+    x3p_percent = x3p_percent * 100,
+    x2p_percent = x2p_percent * 100,
+    e_fg_percent = e_fg_percent * 100,
+    ft_percent = ft_percent * 100
   ) |>
   select(-player, -rk, -tm, -pos, -fg, -fga, -x3p, -x3pa, -x2p, -x2pa, -ft, -fta, -trb) |>
   mutate(id = row_number()) |> # assigning identifier for each player (treating the same player in a different team a different player)
